---
title: "Temperature anomaly in the Canary Islands / World"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: readable
runtime: shiny
---

```{r global, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo    = FALSE,
  error   = FALSE
)
options(scipen = 999)
```

<style>
:root {
  --azul-oscuro: #1B4965;
  --azul-medio:  #2C6E8A;
  --naranja:     #E05C2A;
  --arena:       #8B7355;
  --fondo:       #f4f6f9;
  --blanco:      #ffffff;
  --gris-texto:  #555;
}

body {
  font-family: 'Segoe UI', system-ui, sans-serif;
  font-size: 14px;
  background-color: var(--fondo);
  color: #333;
}

/* ── Navbar ──────────────────────────────────────────────────────────── */
.navbar {
  background-color: var(--azul-oscuro) !important;
  border-bottom: 3px solid var(--naranja);
}
.navbar-brand {
  color: #fff !important;
  font-weight: 700 !important;
  font-size: 15px !important;
  letter-spacing: 0.02em;
}
.navbar-nav > li > a {
  color: #fff !important;
  font-weight: 600;
  letter-spacing: 0.03em;
}
.navbar-nav > li.active > a {
  background-color: var(--naranja) !important;
  color: #fff !important;
  border-radius: 4px;
}

/* ── Panels ──────────────────────────────────────────────────────────── */
.chart-wrapper, .chart-stage {
  background: var(--blanco);
  border-radius: 8px;
  box-shadow: 0 2px 8px rgba(0,0,0,0.07);
}
.chart-title {
  font-size: 12px;
  font-weight: 700;
  color: var(--azul-oscuro);
  padding: 6px 12px 3px;
  border-bottom: 2px solid var(--naranja);
  margin-bottom: 4px;
}

/* ── Sidebar ─────────────────────────────────────────────────────────── */
.section.sidebar {
  background: #e9eef3;
  border-right: 1px solid #cdd6df;
  padding: 14px 12px;
}
.section.sidebar h3 {
  color: var(--azul-oscuro);
  font-size: 13px;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: .06em;
  margin-bottom: 6px;
}

/* ── Value boxes ─────────────────────────────────────────────────────── */
.value-box {
  border-radius: 8px !important;
  box-shadow: 0 2px 8px rgba(0,0,0,0.10) !important;
}
.value-box .value   { font-size: 26px !important; font-weight: 700 !important; }
.value-box .caption { font-size: 12px !important; }

/* ── Interpretive boxes ──────────────────────────────────────────────── */
.caja-discusion {
  background: #fff8f5;
  border-left: 4px solid var(--naranja);
  padding: 10px 14px;
  margin: 8px 0;
  border-radius: 0 6px 6px 0;
  font-size: 13px;
  line-height: 1.6;
}
.caja-hallazgo {
  background: #eef6f8;
  border-left: 4px solid var(--azul-oscuro);
  padding: 10px 14px;
  margin: 8px 0;
  border-radius: 0 6px 6px 0;
  font-size: 13px;
  line-height: 1.6;
}
.caja-metodo {
  background: #f5f0eb;
  border-left: 4px solid var(--arena);
  padding: 10px 14px;
  margin: 8px 0;
  border-radius: 0 6px 6px 0;
  font-size: 13px;
  line-height: 1.6;
}
.caja-conclusion {
  background: #f0f4f8;
  border: 1px solid #c5d5e0;
  border-top: 3px solid var(--azul-oscuro);
  border-radius: 0 0 6px 6px;
  padding: 10px 14px;
  margin-top: 10px;
  font-size: 13px;
  line-height: 1.65;
}
.caja-conclusion .label {
  font-size: 10.5px;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: .06em;
  color: var(--azul-oscuro);
  margin-bottom: 4px;
}

/* ── About compact redesign ───────────────────────────────────────── */
.about-onepage {
  padding: 0px 0px 0px 0px;
}

.about-hero-wrap {
  background: linear-gradient(135deg, #f9fbfd 0%, #eef4f8 100%);
  border: 1px solid #d8e2e9;
  border-top: 4px solid var(--naranja);
  border-radius: 12px;
  padding: 10px 16px 12px 16px;
  box-shadow: 0 2px 6px rgba(0,0,0,0.04);
}

.about-topline {
  font-size: 10.5px;
  font-weight: 800;
  text-transform: uppercase;
  letter-spacing: .12em;
  color: var(--azul-medio);
  margin-bottom: 6px;
}

.about-main-title {
  font-size: 30px;
  line-height: 1.06;
  font-weight: 800;
  color: var(--azul-oscuro);
  margin: 0 0 6px 0;
}

.about-main-subtitle {
  font-size: 14px;
  line-height: 1.5;
  color: #516571;
  max-width: 1050px;
  margin-bottom: 6px;
}

.about-meta-line {
  font-size: 12.5px;
  color: #6b7b86;
  margin-bottom: 12px;
}

.about-highlight-bar {
  background: #fff8f5;
  border-left: 4px solid var(--naranja);
  border-radius: 0 8px 8px 0;
  padding: 7px 11px;
  font-size: 12.6px;
  line-height: 1.55;
  color: #3f4b53;
  margin-bottom: 10px;
}

.about-content-grid {
  display: grid;
  grid-template-columns: 0.84fr 1.16fr;
  gap: 14px;
  align-items: stretch;
}

.about-copy-block {
  background: rgba(255,255,255,0.75);
  border: 1px solid #dde6ec;
  border-radius: 10px;
  padding: 14px 16px;
  display: flex;
  flex-direction: column;
  justify-content: space-between;
}

.about-copy-title {
  font-size: 17px;
  font-weight: 700;
  color: var(--azul-oscuro);
  margin-bottom: 8px;
}

.about-copy-block p {
  font-size: 13px;
  line-height: 1.68;
  color: #3f4e57;
  margin: 0 0 8px 0;
}

.about-copy-block p:last-child {
  margin-bottom: 0;
}

.about-keyfacts {
  margin-top: 10px;
  padding-top: 8px;
  border-top: 1px solid #e4ebf0;
  font-size: 12.5px;
  line-height: 1.6;
  color: #5a6b76;
}

.about-reliability-box {
  margin-top: 10px;
  background: #f9fbfc;
  border: 1px solid #dbe6ed;
  border-radius: 10px;
  padding: 9px 11px;
}

.about-reliability-title {
  font-size: 11.5px;
  font-weight: 800;
  text-transform: uppercase;
  letter-spacing: .07em;
  color: var(--azul-oscuro);
  margin-bottom: 6px;
}

.about-reliability-intro {
  font-size: 12px;
  line-height: 1.5;
  color: #4e5f69;
  margin-bottom: 6px;
}

.about-reliability-list {
  display: grid;
  grid-template-columns: 1fr;
  gap: 3px;
}

.about-reliability-item {
  font-size: 11.8px;
  line-height: 1.42;
  color: #42515a;
  padding-left: 0;
}

.about-reliability-item .tag {
  display: inline-block;
  min-width: 72px;
  font-weight: 700;
  color: var(--azul-oscuro);
}

.about-map-card {
  background: rgba(255,255,255,0.82);
  border: 1px solid #dde6ec;
  border-radius: 10px;
  padding: 10px 10px 8px 10px;
  display: flex;
  flex-direction: column;
  min-height: 100%;
}

.about-map-title {
  font-size: 16px;
  font-weight: 700;
  color: var(--azul-oscuro);
  margin: 2px 2px 10px 2px;
}

.about-map-caption {
  margin-top: 6px;
  font-size: 11.8px;
  line-height: 1.5;
  color: #6a7a86;
}

@media (max-width: 1150px) {
  .about-content-grid {
    grid-template-columns: 1fr;
  }

  .about-main-title {
    font-size: 25px;
  }
}
/* ── KPI badge ───────────────────────────────────────────────────────── */
.kpi-badge {
  display: inline-block;
  padding: 2px 8px;
  border-radius: 12px;
  font-weight: 700;
  font-size: 12px;
}
.kpi-ok  { background: #d4edda; color: #155724; }
.kpi-war { background: #fff3cd; color: #856404; }
.kpi-err { background: #f8d7da; color: #721c24; }

/* ── DT tables ───────────────────────────────────────────────────────── */
.dataTables_wrapper { font-size: 12.5px; }
table.dataTable thead th {
  background: var(--azul-oscuro);
  color: #fff;
  font-weight: 600;
}

/* ── Secondary tabs ──────────────────────────────────────────────────── */
.nav-tabs > li.active > a,
.nav-tabs > li.active > a:hover {
  color: var(--azul-oscuro) !important;
  border-bottom: 3px solid var(--naranja) !important;
  font-weight: 700;
}
.nav-tabs > li > a { color: var(--gris-texto); }
.nav-tabs { margin-bottom: 6px !important; }

/* ── Select input ────────────────────────────────────────────────────── */
.selectize-input {
  border: 1px solid #b5c4cf !important;
  border-radius: 5px !important;
  font-size: 13px !important;
}

/* ── Tabset fill ─────────────────────────────────────────────────────── */
.tabset-fill { height: 100%; }
.tabset-fill .tab-content { height: calc(100% - 52px); padding-top: 4px !important; }
.tabset-fill .tab-pane    { height: 100%; }
.tabset-fill .dataTables_wrapper,
.tabset-fill .dataTables_scroll,
.tabset-fill .dataTables_scrollBody { height: 100% !important; }
.tabset-fill .dataTables_scrollBody { min-height: 280px; }
.nota-bajo-tabset { margin-top: 8px; font-size: 11.5px; }

/* ── Alt models redesign ─────────────────────────────────────────────── */
.finding-box {
  background: #eef6f8;
  border: 1px solid #b8d4e0;
  border-top: 4px solid var(--azul-oscuro);
  border-radius: 0 0 8px 8px;
  padding: 12px 16px;
  font-size: 13px;
  line-height: 1.7;
  margin-top: 6px;
}
.finding-box .finding-title {
  font-size: 11px;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: .07em;
  color: var(--azul-oscuro);
  margin-bottom: 6px;
}
.ratio-pill {
  display: inline-block;
  padding: 1px 8px;
  border-radius: 10px;
  font-weight: 700;
  font-size: 12px;
  margin: 1px 2px;
}
.ratio-accel { background: #fde8df; color: #c0392b; }
.ratio-decel { background: #dff0fb; color: #1B4965; }
.ratio-stable { background: #f0f0f0; color: #555; }

/* ── Trend metric cards ───────────────────────────────────────────── */
.trend-metrics-grid {
  display: grid;
  grid-template-columns: repeat(4, minmax(120px, 1fr));
  gap: 10px;
  margin-top: 10px;
}

.trend-metric-card {
  background: #f8fbfc;
  border: 1px solid #d7e2ea;
  border-top: 3px solid var(--azul-oscuro);
  border-radius: 8px;
  padding: 10px 12px;
  min-height: 78px;
}

.trend-metric-label {
  font-size: 11px;
  font-weight: 800;
  text-transform: uppercase;
  letter-spacing: .06em;
  color: var(--azul-oscuro);
  margin-bottom: 6px;
}

.trend-metric-value {
  font-size: 12.8px;
  line-height: 1.45;
  color: #30424d;
  font-weight: 600;
}

.trend-metric-header {
  font-size: 15px;
  font-weight: 700;
  color: var(--azul-oscuro);
  margin-top: 10px;
  margin-bottom: 2px;
}

@media (max-width: 1200px) {
  .trend-metrics-grid {
    grid-template-columns: repeat(2, minmax(120px, 1fr));
  }
}

@media (max-width: 700px) {
  .trend-metrics-grid {
    grid-template-columns: 1fr;
  }
}
</style>

```{r libs}
library(shiny)
library(flexdashboard)
library(tidyverse)
library(lubridate)
library(tsibble)
library(fpp3)
library(purrr)
library(zoo)
library(ggplot2)
library(plotly)
library(leaflet)
library(broom)
library(lmtest)
library(sandwich)
library(trend)
library(nortest)
library(MASS)
library(tseries)
library(knitr)
library(DT)

source("utilities.R")
```

```{r ns_fix}
select      <- dplyr::select
mutate      <- dplyr::mutate
filter      <- dplyr::filter
group_by    <- dplyr::group_by
summarise   <- dplyr::summarise
selectInput <- shiny::selectInput
```

```{r params}
START_DATE     <- as.Date("1970-01-01")
END_DATE       <- as.Date("2024-12-31")
N_RECENT_YEARS <- 10
CACHE_VERSION  <- "v2"

area_order <- c(
  "GLOBAL", "NORTHERN HEMISPHERE", "SOUTHERN HEMISPHERE",
  "CANARY ISLANDS", "GRAN CANARIA/GANDO", "IZAÑA",
  "S.C. DE TENERIFE", "TENERIFE/LOS RODEOS"
)

MONTH_NAMES     <- c("January","February","March","April","May","June",
                     "July","August","September","October","November","December")
month_levels_en <- tolower(MONTH_NAMES)

# ── AEMET station coordinates (verified from official sources) ───────────────
stations_df <- tibble(
  name    = c("IZAÑA", "TENERIFE/LOS RODEOS", "S.C. DE TENERIFE", "GRAN CANARIA/GANDO"),
  label   = c("Izaña\n(2,373 m)", "Tenerife/Los Rodeos\n(632 m)",
              "S.C. de Tenerife\n(36 m)", "Gran Canaria/Gando\n(24 m)"),
  lat     = c(28.309,  28.478,  28.463,  27.926),
  lon     = c(-16.499, -16.341, -16.255, -15.385),
  alt_m   = c(2373,    632,     36,      24),
  island  = c("Tenerife", "Tenerife", "Tenerife", "Gran Canaria"),
  since   = c("1916", "1944", "1920", "1950")
)
```

```{r helpers}
fmt_century <- function(slope, ic, digits = 2) {
  paste0(
    formatC(round(slope * 100, digits), format = "f", digits = digits),
    " ± ",
    formatC(round(ic * 100, digits), format = "f", digits = digits)
  )
}

ensure_yj_cols <- function(df) {
  yj_cols <- c("RMSE.YJ","SW.YJ","AD.YJ","BP.YJ","lambda.YJ","trend.YJ")
  for (col in yj_cols) if (!col %in% names(df)) df[[col]] <- NA_real_
  df
}

calc_reliability <- function(r2, ic, slope, ad, bp, mk, lb) {
  flags <- c(
    if (!is.na(r2) && r2 < 0.10)                        "⛔R²<0.10",
    if (!is.na(ic) && !is.na(slope) && ic > abs(slope)) "⚠️CI>slope",
    if (!is.na(ad) && ad < 0.05)                        "⚠️AD",
    if (!is.na(bp) && bp < 0.05)                        "⚠️BP",
    if (!is.na(mk) && mk > 0.05)                        "⚠️MK n.s.",
    if (!is.na(lb) && lb < 0.05)                        "⚠️LB"
  )
  if (length(flags) == 0) "✅ OK" else paste(flags, collapse = " ")
}

calc_reliability_num <- function(r2, ic, slope, ad, bp, mk, lb) {
  if (!is.na(r2) && r2 < 0.10) return(2L)
  nfaults <- 0L
  if (!is.na(ic) && !is.na(slope) && ic > abs(slope)) nfaults <- nfaults + 1L
  if (!is.na(ad) && ad < 0.05) nfaults <- nfaults + 1L
  if (!is.na(bp) && bp < 0.05) nfaults <- nfaults + 1L
  if (!is.na(mk) && mk > 0.05) nfaults <- nfaults + 1L
  if (!is.na(lb) && lb < 0.05) nfaults <- nfaults + 1L
  nfaults
}

calc_reliability_label <- function(r2, ic, slope, ad, bp, mk, lb) {
  if (!is.na(r2) && r2 < 0.10)                        return("⛔R²")
  if (!is.na(ic) && !is.na(slope) && ic > abs(slope)) return("⚠️CI")
  if (!is.na(ad) && ad < 0.05)                        return("⚠️AD")
  if (!is.na(bp) && bp < 0.05)                        return("⚠️BP")
  if (!is.na(mk) && mk > 0.05)                        return("⚠️MK")
  if (!is.na(lb) && lb < 0.05)                        return("⚠️LB")
  "✅OK"
}

# ── Plot functions ────────────────────────────────────────────────────────────

plot_heatmap_slope <- function(df) {
  df %>%
    ggplot(aes(x = period, y = AREA, fill = slope_century)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.2f", slope_century)), size = 3.0, colour = "black") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Slope\n(°C/century)") +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 11, base_family = "serif") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
          axis.text.y = element_text(size = 9), panel.grid = element_blank())
}

plot_heatmap_reliability <- function(df) {
  df %>%
    ggplot(aes(x = month_name, y = AREA, fill = reliability_num, label = reliability_label)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(size = 3, colour = "black") +
    scale_fill_gradient(low = "chartreuse1", high = "firebrick1") +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 11, base_family = "serif") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
          axis.text.y = element_text(size = 9),
          panel.grid = element_blank(), legend.position = "none")
}

plot_facet_monthly <- function(data_ts, area_sel) {
  # Convert to tibble and use MONTH_NAMES[month(date)] — locale-independent
  bands_area <- data_ts %>%
    as_tibble() %>%
    filter(!is.na(value), AREA == area_sel) %>%
    mutate(date       = as.Date(date),
           month_name = MONTH_NAMES[month(date)],
           month_name = factor(month_name, levels = MONTH_NAMES)) %>%
    group_by(month_name) %>%
    group_modify(~ {
      mod      <- lm(value ~ date, data = .x)
      pred_ols <- predict(mod, interval = "confidence", level = 0.95)
      V_hac    <- sandwich::vcovHAC(mod)
      X        <- model.matrix(mod)
      se_hac   <- sqrt(rowSums((X %*% V_hac) * X))
      tc       <- qt(0.975, df = mod$df.residual)
      tibble(date = .x$date, value = .x$value, fit = pred_ols[,"fit"],
             ols_lo = pred_ols[,"lwr"], ols_hi = pred_ols[,"upr"],
             hac_lo = pred_ols[,"fit"] - tc * se_hac,
             hac_hi = pred_ols[,"fit"] + tc * se_hac)
    }) %>% ungroup()

  if (nrow(bands_area) == 0)
    return(ggplot() +
             annotate("text", x=0.5, y=0.5, label=paste("No data for:", area_sel),
                      size=5, colour="#1B4965") + theme_void())

  ggplot(bands_area, aes(x = date)) +
    geom_ribbon(aes(ymin = hac_lo, ymax = hac_hi), alpha = 0.30, fill = "red") +
    geom_ribbon(aes(ymin = ols_lo, ymax = ols_hi), alpha = 0.25, fill = "green") +
    geom_point(aes(y = value), size = 1, alpha = 0.5, colour = "grey40") +
    geom_line(aes(y = value), linewidth = 0.3, alpha = 0.5, colour = "grey40") +
    geom_line(aes(y = fit), colour = "blue", linewidth = 0.8) +
    facet_wrap(~ month_name, ncol = 4) +
    labs(title = paste("Anomaly by month —", area_sel),
         subtitle = "Bands: 95% OLS CI (green) vs 95% HAC Newey-West CI (red)",
         x = "", y = "Anomaly (°C)") +
    theme_minimal(base_size = 11, base_family = "serif") +
    theme(strip.text = element_text(face = "bold"), legend.position = "top",
          plot.title = element_text(face = "bold", size = 13))
}
```

```{r load_data}
all_data <- readRDS("temperature_anomaly_2025.rds") %>%
  as_tibble() %>%
  filter(as.Date(date) >= START_DATE, as.Date(date) <= END_DATE)

levels(all_data$AREA)[levels(all_data$AREA) == "LAS PALMAS DE GRAN CANARIA/GANDO"] <- "GRAN CANARIA/GANDO"
levels(all_data$AREA)[levels(all_data$AREA) == "SANTA CRUZ DE TENERIFE"]           <- "S.C. DE TENERIFE"
levels(all_data$AREA)[levels(all_data$AREA) == "NH"]      <- "NORTHERN HEMISPHERE"
levels(all_data$AREA)[levels(all_data$AREA) == "SH"]      <- "SOUTHERN HEMISPHERE"
levels(all_data$AREA)[levels(all_data$AREA) == "CANARIAS"] <- "CANARY ISLANDS"

all_data <- all_data %>%
  mutate(AREA = factor(AREA, levels = area_order)) %>%
  as_tsibble(index = date, key = AREA) %>%
  fill_gaps()

all_data_num <- all_data %>%
  as_tibble() %>%
  mutate(time = as.numeric(date), AREA = factor(AREA, levels = area_order))

all_data_annual <- all_data %>%
  as_tibble() %>%
  mutate(year = year(date), AREA = factor(AREA, levels = area_order)) %>%
  group_by(AREA, year) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(value))

all_data_monthly <- all_data %>%
  as_tibble() %>%
  mutate(year = year(date), month = month(date), AREA = factor(AREA, levels = area_order)) %>%
  filter(!is.na(value))
```

```{r cache_results, include=FALSE}
dir.create("cache/dashboard", recursive = TRUE, showWarnings = FALSE)
cache_file <- paste0("cache/dashboard/results_dashboard_", CACHE_VERSION, ".rds")

if (file.exists(cache_file)) {

  cache_objs <- readRDS(cache_file)
  results_lm               <- cache_objs$results_lm
  results_annual           <- cache_objs$results_annual
  results_monthly          <- cache_objs$results_monthly
  results_monthly2         <- cache_objs$results_monthly2
  table_annual_dashboard   <- cache_objs$table_annual_dashboard
  heatmap_slope_data       <- cache_objs$heatmap_slope_data
  heatmap_reliability_data <- cache_objs$heatmap_reliability_data
  diagnostic_summary       <- cache_objs$diagnostic_summary
  bands_monthly            <- cache_objs$bands_monthly
  bands_annual             <- cache_objs$bands_annual
  acceleration             <- cache_objs$acceleration
  table_acceleration_dashboard <- cache_objs$table_acceleration_dashboard
  best_cp                  <- cache_objs$best_cp
  kpi_canary_annual        <- cache_objs$kpi_canary_annual
kpi_canary_annual        <- cache_objs$kpi_canary_annual
kpi_global_annual        <- cache_objs$kpi_global_annual
kpi_nh_annual            <- cache_objs$kpi_nh_annual
kpi_sh_annual            <- cache_objs$kpi_sh_annual
kpi_n_areas_sig          <- cache_objs$kpi_n_areas_sig
kpi_changepoint          <- cache_objs$kpi_changepoint
kpi_max_improvement      <- cache_objs$kpi_max_improvement

} else {

  results_lm <- all_data_num %>%
    group_by(AREA) %>% filter(!is.na(value)) %>% nest() %>%
    mutate(model    = map(data, ~ lm(value ~ time, data = .x)),
           analysis = map(model, ~ tryCatch(lm.analysis(.x, n_recent = N_RECENT_YEARS * 12L),
                                            error = function(e) tibble(error = conditionMessage(e))))) %>%
    select(AREA, analysis) %>% unnest(analysis) %>%
    mutate(reliability = pmap_chr(list(R2, CI95, lm.slope, AD, BP, MK, LB), calc_reliability)) %>%
    mutate(across(any_of(c("lm.slope","CI95","CI95.HAC","sen.slope","sen.CI95",
                            "sen.recent","sen.recent.CI95","YJN.trend","YJR.trend")), ~ . * 12)) %>%
    ensure_yj_cols() %>% ungroup()

  results_annual <- all_data_annual %>%
    group_by(AREA) %>% nest() %>%
    mutate(model    = map(data, ~ lm(value ~ year, data = .x)),
           analysis = map(model, ~ tryCatch(lm.analysis(.x, n_recent = N_RECENT_YEARS),
                                            error = function(e) tibble(error = conditionMessage(e))))) %>%
    select(AREA, analysis) %>% unnest(analysis) %>%
    mutate(reliability = pmap_chr(list(R2, CI95, lm.slope, AD, BP, MK, LB), calc_reliability)) %>%
    ensure_yj_cols() %>% ungroup()

  results_monthly <- all_data_monthly %>%
    group_by(AREA, month) %>% nest() %>%
    mutate(model    = map(data, ~ lm(value ~ year, data = .x)),
           analysis = map(model, ~ tryCatch(lm.analysis(.x, n_recent = N_RECENT_YEARS),
                                            error = function(e) tibble(error = conditionMessage(e))))) %>%
    select(AREA, month, analysis) %>% unnest(analysis) %>%
    mutate(month_name  = factor(MONTH_NAMES[month], levels = MONTH_NAMES),
           reliability = pmap_chr(list(R2, CI95, lm.slope, AD, BP, MK, LB), calc_reliability)) %>%
    arrange(AREA, month) %>% ensure_yj_cols() %>% ungroup()

  results_monthly2 <- results_monthly %>%
    mutate(reliability_num   = pmap_int(list(R2, CI95, lm.slope, AD, BP, MK, LB), calc_reliability_num),
           reliability_label = pmap_chr(list(R2, CI95, lm.slope, AD, BP, MK, LB), calc_reliability_label))

  table_annual_dashboard <- results_annual %>%
    transmute(AREA, R2 = round(R2, 3),
              `OLS ± CI95.HAC` = fmt_century(lm.slope, CI95.HAC),
              `Sen ± CI95`     = fmt_century(sen.slope, sen.CI95),
              reliability)

  heatmap_slope_data <- results_monthly %>%
    select(AREA, month_name, lm.slope) %>%
    rename(period = month_name) %>%
    mutate(AREA = factor(AREA, levels = area_order),
           period = factor(period, levels = MONTH_NAMES),
           slope_century = lm.slope * 100) %>%
    arrange(AREA, period)

  heatmap_reliability_data <- results_monthly2 %>%
    select(AREA, month_name, reliability_num, reliability_label) %>%
    mutate(AREA = factor(AREA, levels = area_order),
           month_name = factor(month_name, levels = MONTH_NAMES)) %>%
    arrange(AREA, month_name)

  diagnostic_summary <- bind_rows(
    results_lm      %>% mutate(scale = "Full monthly"),
    results_annual  %>% mutate(scale = "Annual"),
    results_monthly %>% mutate(scale = "By month")
  ) %>%
    summarise(.by = scale,
      `AD > 0.05 (%)` = round(mean(AD > 0.05, na.rm=TRUE)*100, 1),
      `BP > 0.05 (%)` = round(mean(BP > 0.05, na.rm=TRUE)*100, 1),
      `LB > 0.05 (%)` = round(mean(LB > 0.05, na.rm=TRUE)*100, 1),
      `MK < 0.05 (%)` = round(mean(MK < 0.05, na.rm=TRUE)*100, 1),
      `Mean R²`                 = round(mean(R2, na.rm=TRUE), 3),
      `Mean slope (°C/century)` = round(mean(lm.slope, na.rm=TRUE)*100, 2),
      `CI95 (°C/century)`       = round(mean(CI95, na.rm=TRUE)*100, 2),
      `CI95.HAC (°C/century)`   = round(mean(CI95.HAC, na.rm=TRUE)*100, 2)) %>%
    mutate(scale = factor(scale, levels = c("Full monthly","Annual","By month"))) %>%
    arrange(scale)

  bands_monthly <- all_data %>%
    as_tibble() %>%
    mutate(date = as.Date(date), AREA = factor(AREA, levels = area_order)) %>%
    filter(!is.na(value)) %>%
    group_by(AREA) %>%
    group_modify(~ {
      mod      <- lm(value ~ date, data = .x)
      pred_ols <- predict(mod, interval = "confidence", level = 0.95)
      V_hac    <- sandwich::vcovHAC(mod); X <- model.matrix(mod)
      se_hac   <- sqrt(rowSums((X %*% V_hac) * X))
      tc       <- qt(0.975, df = mod$df.residual)
      tibble(date = .x$date, value = .x$value, fit = pred_ols[,"fit"],
             ols_lo = pred_ols[,"lwr"], ols_hi = pred_ols[,"upr"],
             hac_lo = pred_ols[,"fit"] - tc*se_hac,
             hac_hi = pred_ols[,"fit"] + tc*se_hac)
    }) %>% ungroup()

  bands_annual <- all_data_annual %>%
    group_by(AREA) %>%
    group_modify(~ {
      mod      <- lm(value ~ year, data = .x)
      pred_ols <- predict(mod, interval = "confidence", level = 0.95)
      V_hac    <- sandwich::vcovHAC(mod); X <- model.matrix(mod)
      se_hac   <- sqrt(rowSums((X %*% V_hac) * X))
      tc       <- qt(0.975, df = mod$df.residual)
      tibble(year = .x$year, value = .x$value, fit = pred_ols[,"fit"],
             ols_lo = pred_ols[,"lwr"], ols_hi = pred_ols[,"upr"],
             hac_lo = pred_ols[,"fit"] - tc*se_hac,
             hac_hi = pred_ols[,"fit"] + tc*se_hac)
    }) %>% ungroup() %>%
    mutate(date = as.Date(paste0(year, "-01-01")))

  acceleration                 <- tibble()
  table_acceleration_dashboard <- tibble()
  best_cp             <- as.Date(NA)
  kpi_changepoint     <- "Not computed"
  kpi_max_improvement <- NA_real_

  try({
    optimise_yj_rmse <- function(x, y, lambda_range = c(-2, 3)) {
      m <- mean(y); s <- sd(y); ys <- (y - m) / s
      optimize(function(lambda) {
        yj <- yeo.johnson(ys, lambda); fit <- lm(yj ~ x)
        rt <- m + s * yeo.johnson.inverse(fitted(fit), lambda)
        sum((y - rt)^2)
      }, interval = lambda_range)$minimum
    }

    acceleration <- all_data_num %>%
      filter(!is.na(value)) %>%
      group_by(AREA) %>%
      group_modify(~ {
        y <- .x$value; x <- .x$time; n <- length(y)
        mod        <- lm(value ~ time, data = .x)
        slope_hist <- coef(mod)[2]
        ci95_hac   <- qnorm(0.975) * sqrt(diag(sandwich::vcovHAC(mod)))[2]
        lambda     <- optimise_yj_rmse(x, y)
        m <- mean(y); s <- sd(y); ys <- (y - m) / s
        yj     <- yeo.johnson(ys, lambda); fit_yj <- lm(yj ~ x)
        yt2    <- m + s * yeo.johnson.inverse(fitted(fit_yj), lambda)
        rmse_ols        <- sqrt(mean((y - fitted(mod))^2))
        rmse_yj         <- sqrt(mean((y - yt2)^2))
        n_rec           <- min(N_RECENT_YEARS * 12L, max(5L, n %/% 3L))
        yj_slope_recent <- coef(lm(tail(yt2, n_rec) ~ tail(x, n_rec)))[2]
        tibble(slope_hist = slope_hist*12, ci95_hac = ci95_hac*12,
               yj_recent = yj_slope_recent*12, lambda_yj = round(lambda,2),
               rmse_ols = round(rmse_ols,4), rmse_yj = round(rmse_yj,4))
      }) %>% ungroup()

    n_dates    <- length(unique(all_data$date))
    margin     <- 0.15
    candidates <- sort(unique(all_data$date))[ceiling(n_dates*margin):floor(n_dates*(1-margin))]

    best_cp <- map_df(candidates, function(cp) {
      fit <- all_data %>%
        model(m = TSLM(value ~ trend() + I(pmax(0, as.numeric(date) - as.numeric(cp)))))
      tibble(cp = cp, mean_rmse = mean(accuracy(fit)$RMSE, na.rm = TRUE))
    }) %>% slice_min(mean_rmse, n = 1, with_ties = FALSE) %>% pull(cp)

    model_cp  <- all_data %>%
      model(m = TSLM(value ~ trend() + I(pmax(0, as.numeric(date) - as.numeric(best_cp)))))
    metrics_cp <- accuracy(model_cp) %>% select(AREA, RMSE)
    coefs_cp   <- tidy(model_cp) %>%
      group_by(AREA) %>%
      summarise(slope1 = estimate[term == "trend()"],
                slope2 = slope1 + estimate[grepl("pmax", term)], .groups = "drop")

    result_cp <- coefs_cp %>%
      left_join(metrics_cp, by = "AREA") %>%
      mutate(changepoint = best_cp) %>%
      rename(rmse_cp = RMSE, slope2_cp = slope2) %>%
      select(AREA, changepoint, rmse_cp, slope2_cp)

    acceleration <- acceleration %>% left_join(result_cp, by = "AREA")

    table_acceleration_dashboard <- acceleration %>%
      mutate(AREA = factor(AREA, levels = area_order)) %>%
      arrange(AREA) %>%
      transmute(
        AREA,
        `RMSE (OLS)` = rmse_ols, `RMSE (YJ)` = rmse_yj,
        `RMSE (CP)`  = round(rmse_cp, 4),
        `YJ improv. (%)` = round((1 - rmse_yj/rmse_ols)*100, 2),
        `CP improv. (%)` = round((1 - rmse_cp/rmse_ols)*100, 2),
        `Historical ± CI95.HAC (°C/century)` = fmt_century(slope_hist, ci95_hac),
        `YJ recent (°C/century)`  = round(yj_recent*100, 2),
        `CP slope 2 (°C/century)` = round(slope2_cp*12*100, 2),
        `Changepoint`             = format(changepoint, "%Y-%m"),
        `YJ/hist. ratio`          = round(yj_recent/slope_hist, 2),
        `CP/hist. ratio`          = round(slope2_cp*12/slope_hist, 2)
      )

    kpi_changepoint     <- format(best_cp, "%Y-%m")
    kpi_max_improvement <- table_acceleration_dashboard %>%
      summarise(m = max(c(`CP improv. (%)`, `YJ improv. (%)`), na.rm=TRUE)) %>% pull(m)
  }, silent = TRUE)

  kpi_canary_annual <- results_annual %>% filter(AREA == "CANARY ISLANDS")      %>% pull(lm.slope) %>% .[1]
kpi_global_annual <- results_annual %>% filter(AREA == "GLOBAL")              %>% pull(lm.slope) %>% .[1]
kpi_nh_annual     <- results_annual %>% filter(AREA == "NORTHERN HEMISPHERE") %>% pull(lm.slope) %>% .[1]
kpi_sh_annual     <- results_annual %>% filter(AREA == "SOUTHERN HEMISPHERE") %>% pull(lm.slope) %>% .[1]
kpi_n_areas_sig   <- results_annual %>% filter(MK < 0.05) %>% nrow()

  saveRDS(list(
    results_lm = results_lm, results_annual = results_annual,
    results_monthly = results_monthly, results_monthly2 = results_monthly2,
    table_annual_dashboard = table_annual_dashboard,
    heatmap_slope_data = heatmap_slope_data, heatmap_reliability_data = heatmap_reliability_data,
    diagnostic_summary = diagnostic_summary, bands_monthly = bands_monthly,
    bands_annual = bands_annual, acceleration = acceleration,
    table_acceleration_dashboard = table_acceleration_dashboard,
    best_cp = best_cp, kpi_canary_annual = kpi_canary_annual,
kpi_global_annual = kpi_global_annual, kpi_nh_annual = kpi_nh_annual,
kpi_sh_annual = kpi_sh_annual, kpi_n_areas_sig = kpi_n_areas_sig,
kpi_changepoint = kpi_changepoint, kpi_max_improvement = kpi_max_improvement
  ), cache_file)
}
```


<!-- =====================================================================
     TAB 0 – ABOUT
     ===================================================================== -->

About
=====================================

Row {data-height=640}
-------------------------------------

### {.no-title}

```{r}
htmltools::tagList(

  htmltools::HTML('
  <div class="about-onepage">
    <div class="about-hero-wrap">

      <div class="about-main-title">
        Temperature Anomaly Dashboard · Canary Islands & Global Context (1970–2024)
      </div>

      <div class="about-main-subtitle">
        Long-term warming trends in the Canary Islands are analysed in a global context
        using monthly temperature anomaly series and robust statistical diagnostics.
      </div>

      <div class="about-highlight-bar">
        <b>Main finding.</b> All analysed areas show significant warming, with the Canary Islands
        regional trend broadly comparable to the Northern Hemisphere and clear seasonal differences across stations.
      </div>

      <div class="about-content-grid">
        <div class="about-copy-block">
          <div>
            <div class="about-copy-title">Project overview</div>

            <p>
  This dashboard analyses temperature anomaly trends for the global mean,
  both hemispheres, four long-record AEMET stations, and a regional Canary Islands mean,
  combining monthly, annual, and month-specific perspectives to capture both
  <b>long-term warming intensity</b> and <b>seasonal structure</b> over the period <b>1970–2024</b>.
</p>

<p>
  The analysis uses data from <b>HadCRUT5</b> and <b>AEMET</b> and is supported by
  <b>HAC confidence intervals</b>, <b>Sen’s slope</b>, <b>Mann-Kendall testing</b>,
  and standard residual diagnostics, while alternative models are included only
  to assess <b>recent-trend changes</b> and departures from linearity.
</p>

            <div class="about-reliability-box">
              <div class="about-reliability-title">Reliability indicators</div>

              <div class="about-reliability-intro">
                Each model receives an automatic diagnostic label. When no warning is triggered,
                the model is classified as <b>✅ OK</b>.
              </div>

              <div class="about-reliability-list">
                <div class="about-reliability-item">
                  <span class="tag">⛔ R²&lt;0.10</span>
                  trend signal dominated by noise.
                </div>
                <div class="about-reliability-item">
                  <span class="tag">⚠️ CI&gt;slope</span>
                  trend not clearly distinguishable from zero.
                </div>
                <div class="about-reliability-item">
                  <span class="tag">⚠️ AD</span>
                  residual normality rejected.
                </div>
                <div class="about-reliability-item">
                  <span class="tag">⚠️ BP</span>
                  heteroscedasticity detected.
                </div>
                <div class="about-reliability-item">
                  <span class="tag">⚠️ LB</span>
                  residual autocorrelation detected.
                </div>
                <div class="about-reliability-item">
                  <span class="tag">⚠️ MK n.s.</span>
                  Mann-Kendall does not detect a significant monotonic trend.
                </div>
              </div>
            </div>
          </div>
        </div>

        <div class="about-map-card">
          <div class="about-map-title">AEMET stations used in the study</div>
  '),

  leaflet(stations_df, height = 400) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(lng = -15.9, lat = 28.2, zoom = 7) %>%
    addCircleMarkers(
      lng = ~lon,
      lat = ~lat,
      radius = ~pmax(5, sqrt(alt_m) * 0.22),
      color = "#1B4965",
      fillColor = "#E05C2A",
      fillOpacity = 0.72,
      weight = 1.4,
      popup = ~paste0(
        "<b>", name, "</b><br>",
        "Island: ", island, "<br>",
        "Altitude: ", format(alt_m, big.mark=","), " m a.s.l.<br>",
        "Records since: ", since
      ),
      label = ~name
    ) %>%
    addLegend(
      position = "bottomright",
      colors = "#E05C2A",
      labels = "AEMET station",
      title = "Study network",
      opacity = 0.85
    ),

  htmltools::HTML('
          <div class="about-map-caption">
            Four long-term AEMET stations selected to represent contrasting coastal and high-altitude environments across the archipelago.
          </div>
        </div>
      </div>

    </div>
  </div>
  ')
)
```


<!-- =====================================================================
     TAB 1 – EXECUTIVE SUMMARY
     ===================================================================== -->

Executive Summary
=====================================

Row {data-height=130}
-------------------------------------

### 🌴 Canary Islands

```{r}
valueBox(value   = paste0(round(kpi_canary_annual * 100, 2), " °C/century"),
         caption = "Annual slope · Canary Islands",
         icon    = "fa-thermometer-half", color = "#1B4965")
```

### 🌍 Global

```{r}
valueBox(value   = paste0(round(kpi_global_annual * 100, 2), " °C/century"),
         caption = "Annual slope · Global",
         icon    = "fa-globe", color = "#2C6E8A")
```

### 🧭 Northern Hemisphere

```{r}
valueBox(value   = paste0(round(kpi_nh_annual * 100, 2), " °C/century"),
         caption = "Annual slope · Northern Hemisphere",
         icon    = "fa-line-chart", color = "#8B7355")
```

### 🧭 Southern Hemisphere

```{r}
valueBox(
  value   = paste0(round(kpi_sh_annual * 100, 2), " °C/century"),
  caption = "Annual slope · Southern Hemisphere",
  icon    = "fa-arrow-down",
  color   = "#E05C2A"
)
```

Row
-------------------------------------

### Annual slope comparison by area {data-width=600}

```{r}
bars_data <- results_annual %>%
  mutate(AREA = factor(AREA, levels = rev(area_order)),
         slope_century = lm.slope * 100, ci_century = CI95.HAC * 100,
         tooltip_txt = paste0(
           "<b>", AREA, "</b><br>",
           "OLS slope: <b>", round(lm.slope*100,2), " °C/century</b><br>",
           "HAC CI95: ± ", round(CI95.HAC*100,2), "<br>",
           "R²: ", round(R2,3), "<br>",
           "Reliability: ", reliability))

plot_ly(data=bars_data, x=~slope_century, y=~AREA, type="bar", orientation="h",
        error_x=list(type="data", array=~ci_century, color="#E05C2A", thickness=1.5),
        marker=list(color="#1B4965"), hovertext=~tooltip_txt, hoverinfo="text") %>%
  layout(title  = list(text="Annual slope (°C/century) · OLS with HAC CI95", font=list(size=14)),
         xaxis  = list(title="°C/century", zeroline=TRUE, zerolinecolor="#aaa"),
         yaxis  = list(title=""),
         margin = list(l=180, r=20, t=50, b=40),
         plot_bgcolor="rgba(0,0,0,0)", paper_bgcolor="rgba(0,0,0,0)") %>%
  config(displayModeBar=FALSE)
```

### Quick diagnostics {data-width=400}

```{r}
### Quick diagnostics {data-width=400}

```{r}
diagnostic_summary_t <- diagnostic_summary %>%
  mutate(scale = as.character(scale)) %>%
  tibble::column_to_rownames("scale") %>%
  t() %>%
  as.data.frame(check.names = FALSE) %>%
  tibble::rownames_to_column("Metric")

div(class = "tabset-fill", tagList(
  tabsetPanel(
    tabPanel("📋 Summary table",
      tagList(
        DT::datatable(
          table_annual_dashboard,
          rownames = FALSE,
          class = "compact stripe",
          options = list(
            paging = FALSE, searching = FALSE, info = FALSE,
            ordering = TRUE, autoWidth = TRUE, scrollX = FALSE
          )
        ),
        tags$div(
          class = "caja-metodo nota-bajo-tabset",
          HTML("<b>HAC CI:</b> corrected for autocorrelation and heteroscedasticity (Newey-West).
               <b>Sen slope:</b> robust Theil-Sen estimator, resistant to outliers.
               Full indicator guide available in the <b>About</b> tab.")
        )
      )
    ),
    tabPanel("🔬 Diagnostics",
      DT::datatable(
        diagnostic_summary_t,
        rownames = FALSE,
        class = "compact stripe",
        options = list(
          paging = FALSE, searching = FALSE, info = FALSE,
          ordering = FALSE, autoWidth = TRUE, scrollX = TRUE
        )
      ) %>%
        DT::formatStyle("Metric", fontWeight = "bold")
    )
  )
))
```


<!-- =====================================================================
     TAB 2 – TRENDS
     ===================================================================== -->

Trends
=====================================

Sidebar {.sidebar}
-------------------------------------

### Area selection

```{r}
selectInput(inputId = "area_trend", label = "Select area:",
            choices = area_order, selected = "CANARY ISLANDS")

renderUI({
  req(input$area_trend)

  row_m <- results_lm %>%
    filter(AREA == input$area_trend) %>%
    slice(1)

  row_a <- results_annual %>%
    filter(AREA == input$area_trend) %>%
    slice(1)

  dif <- round((row_m$lm.slope - row_a$lm.slope) * 100, 2)
  dir <- if (dif > 0) "higher" else "lower"

  htmltools::HTML(paste0(
    "<div class='caja-conclusion' style='margin-top:8px'>",
    "<div class='label'>🔎 Summary</div>",
    "Monthly slope is <b>", dir, "</b> than annual by <b>", abs(dif), " °C/century</b>. ",
    "Notable differences suggest seasonality influences the overall estimate.",
    "</div>",
    "<div class='caja-metodo' style='margin-top:8px; font-size:11.5px'>",
    "<b>HAC CI:</b> corrects autocorrelation/heteroscedasticity.<br>",
    "<b>Sen:</b> robust estimator (Theil-Sen).",
    "</div>"
  ))
})
```

Row
-------------------------------------

### {.no-title}

```{r}
tabsetPanel(

  tabPanel("📅 Monthly series",

    renderUI({
      req(input$area_trend)

      row_m <- results_lm %>%
        filter(AREA == input$area_trend) %>%
        slice(1)

      badge_class <- function(rel) {
        if (grepl("⛔", rel)) "kpi-err"
        else if (grepl("⚠️", rel)) "kpi-war"
        else "kpi-ok"
      }

      htmltools::HTML(paste0(
        "<div class='caja-hallazgo' style='margin-bottom:10px'>",
        "<b>R²:</b> ", round(row_m$R2, 3),
        " &nbsp;&nbsp;|&nbsp;&nbsp; ",
        "<b>OLS HAC:</b> ", fmt_century(row_m$lm.slope, row_m$CI95.HAC), " °C/century",
        " &nbsp;&nbsp;|&nbsp;&nbsp; ",
        "<b>Sen:</b> ", fmt_century(row_m$sen.slope, row_m$sen.CI95), " °C/century",
        " &nbsp;&nbsp;|&nbsp;&nbsp; ",
        "<b>Status:</b> <span class='kpi-badge ", badge_class(row_m$reliability), "'>",
        row_m$reliability, "</span>",
        "</div>"
      ))
    }),

    renderPlotly({
      req(input$area_trend)
      d <- bands_monthly %>% filter(AREA == input$area_trend)

      plot_ly(d, x = ~date) %>%
        add_ribbons(
          ymin = ~hac_lo, ymax = ~hac_hi,
          fillcolor = "rgba(224,92,42,0.15)",
          line = list(color = "transparent"),
          name = "HAC CI95"
        ) %>%
        add_lines(
          y = ~value,
          line = list(color = "rgba(120,120,120,0.6)", width = 0.8),
          name = "Observed anomaly"
        ) %>%
        add_lines(
          y = ~fit,
          line = list(color = "#1B4965", width = 2.2),
          name = "OLS trend"
        ) %>%
        layout(
          title = list(text = paste("Monthly series:", input$area_trend), font = list(size = 14)),
          xaxis = list(title = ""),
          yaxis = list(title = "Anomaly (°C)"),
          legend = list(orientation = "h", y = -0.18),
          hovermode = "x unified",
          plot_bgcolor = "rgba(0,0,0,0)",
          paper_bgcolor = "rgba(0,0,0,0)"
        ) %>%
        config(displayModeBar = FALSE)
    })
  ),

  tabPanel("📆 Annual mean",

    renderUI({
      req(input$area_trend)

      row_a <- results_annual %>%
        filter(AREA == input$area_trend) %>%
        slice(1)

      badge_class <- function(rel) {
        if (grepl("⛔", rel)) "kpi-err"
        else if (grepl("⚠️", rel)) "kpi-war"
        else "kpi-ok"
      }

      htmltools::HTML(paste0(
        "<div class='caja-hallazgo' style='margin-bottom:10px'>",
        "<b>R²:</b> ", round(row_a$R2, 3),
        " &nbsp;&nbsp;|&nbsp;&nbsp; ",
        "<b>OLS HAC:</b> ", fmt_century(row_a$lm.slope, row_a$CI95.HAC), " °C/century",
        " &nbsp;&nbsp;|&nbsp;&nbsp; ",
        "<b>Sen:</b> ", fmt_century(row_a$sen.slope, row_a$sen.CI95), " °C/century",
        " &nbsp;&nbsp;|&nbsp;&nbsp; ",
        "<b>Status:</b> <span class='kpi-badge ", badge_class(row_a$reliability), "'>",
        row_a$reliability, "</span>",
        "</div>"
      ))
    }),

    renderPlotly({
      req(input$area_trend)
      d <- bands_annual %>%
        filter(AREA == input$area_trend) %>%
        mutate(date = as.Date(paste0(year, "-01-01")))

      plot_ly(d, x = ~date) %>%
        add_ribbons(
          ymin = ~hac_lo, ymax = ~hac_hi,
          fillcolor = "rgba(224,92,42,0.15)",
          line = list(color = "transparent"),
          name = "HAC CI95"
        ) %>%
        add_lines(
          y = ~value,
          line = list(color = "rgba(100,100,100,0.65)", width = 1),
          name = "Observed annual mean"
        ) %>%
        add_markers(
          y = ~value,
          marker = list(color = "#8B7355", size = 5),
          name = "",
          showlegend = FALSE
        ) %>%
        add_lines(
          y = ~fit,
          line = list(color = "#1B4965", width = 2.5),
          name = "OLS trend"
        ) %>%
        layout(
          title = list(text = paste("Annual mean:", input$area_trend), font = list(size = 14)),
          xaxis = list(title = ""),
          yaxis = list(title = "Annual mean anomaly (°C)"),
          legend = list(orientation = "h", y = -0.18),
          hovermode = "x unified",
          plot_bgcolor = "rgba(0,0,0,0)",
          paper_bgcolor = "rgba(0,0,0,0)"
        ) %>%
        config(displayModeBar = FALSE)
    })
  )
)
```

<!-- =====================================================================
     TAB 3 – SEASONALITY
     ===================================================================== -->

Seasonality
=====================================

Sidebar {.sidebar}
-------------------------------------

### Area selection

```{r}
selectInput(inputId="area_month", label="Select area:",
            choices=area_order, selected="CANARY ISLANDS")
```

```{r}
htmltools::HTML("
  <div class='caja-metodo'>
    <b>How to read this tab</b><br><br>
    The two heatmaps summarise <b>warming intensity</b> and <b>statistical
    reliability</b> for each month and area.<br><br>
    Below: faceted evolution by month for the selected area.
  </div>
  <div class='caja-conclusion' style='margin-top:8px'>
    <div class='label'>🔎 Summary</div>
    Warming is not uniform throughout the year. Canarian stations peak
    in spring–summer; the Northern Hemisphere peaks in winter.
    7 of 96 monthly models show ⛔R² &lt; 0.10, concentrated at
    Izaña and Tenerife/Los Rodeos in winter months.
  </div>
")
```

Row
-------------------------------------

### Warming intensity heatmap (°C/century)

```{r, out.width="100%", fig.asp=0.4}
plot_heatmap_slope(heatmap_slope_data)
```

### Statistical reliability heatmap

```{r, out.width="100%", fig.asp=0.4}
plot_heatmap_reliability(heatmap_reliability_data)
```

Row
-------------------------------------
### Monthly anomaly evolution · selected area

```{r}
renderPlot({
  req(input$area_month)
  plot_facet_monthly(all_data, input$area_month)
})
```


<!-- =====================================================================
     TAB 4 – ALTERNATIVE MODELS  (redesigned)
     ===================================================================== -->

Alternative Models
=====================================

```{r alt_flags, include=FALSE}
has_yj <- nrow(acceleration) > 0 &&
  all(c("AREA","slope_hist","yj_recent","rmse_ols","rmse_yj") %in% names(acceleration))
has_cp <- nrow(acceleration) > 0 &&
  all(c("slope2_cp","rmse_cp","changepoint") %in% names(acceleration))
```

Row {data-height=90}
-------------------------------------

### {.no-title}

```{r}
# Contextual banner — always visible regardless of whether models ran
htmltools::HTML(paste0(
  "<div style='padding:10px 16px; background:#f0f4f8; border-bottom:2px solid #1B4965;",
  "font-size:13px; line-height:1.6;'>",
  "<b>Why alternative models?</b> &nbsp;OLS assumes a <em>constant linear trend</em>. ",
  "Yeo-Johnson and segmented regression (changepoint) allow the trend to curve or break. ",
  "If their RMSE improvement over OLS is small (&lt;2%), linear regression already captures the underlying trend well — ",
  "which is exactly what this study finds. Both methods also provide an estimate of the <em>recent rate of change</em>, ",
  "allowing comparison with the long-run historical slope.",
  if (has_cp) paste0(" &nbsp;|&nbsp; <b>Optimal changepoint:</b> ", kpi_changepoint,
                     " &nbsp;|&nbsp; <b>Max. RMSE improvement:</b> ",
                     round(kpi_max_improvement, 2), "%") else "",
  "</div>"
))
```

Row {data-height=390}
-------------------------------------

### Historical vs. recent slope by area {data-width=560}

```{r, fig.asp=0.72}
if (has_yj) {
  plot_data <- acceleration %>%
    mutate(
      AREA = factor(AREA, levels = rev(area_order)),
      `Historical OLS`          = slope_hist * 100,
      `Yeo-Johnson (recent)`    = yj_recent * 100,
      `Changepoint (segment 2)` = if ("slope2_cp" %in% names(.)) slope2_cp * 12 * 100 else NA_real_
    ) %>%
    select(AREA, `Historical OLS`, `Yeo-Johnson (recent)`, `Changepoint (segment 2)`) %>%
    pivot_longer(cols = -AREA, names_to = "method", values_to = "slope") %>%
    filter(!is.na(slope))

  ggplot(plot_data, aes(x = slope, y = AREA, fill = method)) +
    geom_col(position = position_dodge(width = 0.72), width = 0.62) +
    geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey50") +
    scale_fill_manual(values = c(
      "Historical OLS"          = "#1B4965",
      "Yeo-Johnson (recent)"    = "#E05C2A",
      "Changepoint (segment 2)" = "#2C6E8A"
    )) +
    labs(title    = "Historical slope vs. recent estimates (°C/century)",
         subtitle = "Bars right of Historical OLS → acceleration · left → deceleration",
         x = "Slope (°C/century)", y = NULL, fill = NULL) +
    theme_minimal(base_size = 12, base_family = "serif") +
    theme(panel.grid.minor  = element_blank(),
          axis.text.y       = element_text(face = "bold", size = 10),
          plot.title        = element_text(face = "bold", size = 13),
          plot.subtitle     = element_text(size = 10.5, colour = "#555"),
          legend.position   = "top")
} else {
  htmltools::HTML("<div class='caja-discusion'>Alternative models not available in this run.</div>")
}
```

### Interpretation & ratios {data-width=440}

```{r}
if (has_yj) {

  # Build ratio pills per area
  ratio_pills <- if (has_cp) {
    acceleration %>%
      mutate(AREA = factor(AREA, levels = area_order),
             ratio_yj = round(yj_recent / slope_hist, 2),
             ratio_cp = round(slope2_cp * 12 / slope_hist, 2)) %>%
      arrange(AREA) %>%
      pmap_chr(function(AREA, ratio_yj, ratio_cp, ...) {
        pill_class <- function(r) {
          if (is.na(r)) return("")
          if (r > 1.05) "ratio-accel" else if (r < 0.95) "ratio-decel" else "ratio-stable"
        }
        paste0(
          "<tr>",
          "<td style='padding:3px 8px; font-weight:600; font-size:12px'>", AREA, "</td>",
          "<td style='padding:3px 6px; text-align:center'>",
          "<span class='ratio-pill ", pill_class(ratio_yj), "'>", ratio_yj, "</span></td>",
          "<td style='padding:3px 6px; text-align:center'>",
          "<span class='ratio-pill ", pill_class(ratio_cp), "'>", ratio_cp, "</span></td>",
          "</tr>"
        )
      }) %>% paste(collapse = "")
  } else {
    acceleration %>%
      mutate(AREA = factor(AREA, levels = area_order),
             ratio_yj = round(yj_recent / slope_hist, 2)) %>%
      arrange(AREA) %>%
      pmap_chr(function(AREA, ratio_yj, ...) {
        pill_class <- function(r) if (is.na(r)) "" else if (r > 1.05) "ratio-accel" else if (r < 0.95) "ratio-decel" else "ratio-stable"
        paste0("<tr><td style='padding:3px 8px; font-weight:600; font-size:12px'>", AREA, "</td>",
               "<td style='padding:3px 6px; text-align:center'>",
               "<span class='ratio-pill ", pill_class(ratio_yj), "'>", ratio_yj, "</span></td></tr>")
      }) %>% paste(collapse="")
  }

  header_cp <- if (has_cp) "<th style='padding:4px 6px; text-align:center'>CP/hist.</th>" else ""

  htmltools::HTML(paste0(
    "<div style='padding:6px 4px; overflow-y:auto; height:100%'>",

    "<div class='finding-box'>",
    "<div class='finding-title'>📌 Key finding</div>",
    "RMSE improvement over OLS is <b>&lt;2% in all areas</b>, confirming that simple linear
     regression adequately captures the long-run trend. The non-linearity present
     is too weak to be meaningfully exploited by more flexible models.",
    "</div>",

    "<div class='caja-discusion' style='margin-top:10px'>",
    "<b>Global series → moderate acceleration</b><br>",
    "Yeo-Johnson ratios: 1.27 (Global), 1.28 (N. Hemisphere).<br>",
    "Changepoint ratios: 1.07 (Global), 1.08 (N. Hemisphere).<br>",
    "Both methods agree: warming is accelerating at planetary scale.",
    "</div>",

    "<div class='caja-hallazgo' style='margin-top:8px'>",
    "<b>Canarian stations → possible deceleration</b><br>",
    "Yeo-Johnson ratios: 0.88–1.01 · Changepoint ratios: 0.58–0.96.<br>",
    "Signal is weak (RMSE &lt;2%) but consistent across both methods.
     Could reflect trade wind changes or oceanic upwelling effects.",
    "</div>",

    "<div style='margin-top:12px'>",
    "<b style='font-size:12px; color:#1B4965'>Ratio table</b>
     <span style='font-size:11px; color:#888'>&nbsp;(>1.05 = ",
    "<span class='ratio-pill ratio-accel'>accel</span> &nbsp;",
    "<1.05 = <span class='ratio-pill ratio-decel'>decel</span> &nbsp;",
    "≈1 = <span class='ratio-pill ratio-stable'>stable</span>)",
    "</span>",
    "<table style='width:100%; border-collapse:collapse; margin-top:6px'>",
    "<thead><tr style='background:#1B4965; color:#fff'>",
    "<th style='padding:4px 8px; text-align:left'>Area</th>",
    "<th style='padding:4px 6px; text-align:center'>YJ/hist.</th>",
    header_cp,
    "</tr></thead><tbody>",
    ratio_pills,
    "</tbody></table></div>",

    "<div class='caja-metodo' style='margin-top:10px; font-size:11.5px'>",
    "<b>Ratio interpretation:</b> recent slope / historical slope.
     A ratio of 1.27 means the recent rate is 27% higher than the historical average.",
    "</div>",

    "</div>"
  ))
} else {
  htmltools::HTML("<div class='caja-discusion'>No summary available for alternative models.</div>")
}
```

Row {data-height=340}
-------------------------------------

### RMSE goodness-of-fit by method {data-width=560}

```{r, fig.asp=0.55}
if (has_yj) {
  acceleration %>%
    mutate(AREA = factor(AREA, levels = area_order)) %>%
    select(AREA, rmse_ols, rmse_yj, any_of("rmse_cp")) %>%
    pivot_longer(cols = -AREA, names_to = "method", values_to = "rmse") %>%
    filter(!is.na(rmse)) %>%
    mutate(method = factor(method,
      levels = c("rmse_ols","rmse_yj","rmse_cp"),
      labels = c("OLS","Yeo-Johnson","Changepoint"))) %>%
    ggplot(aes(x = AREA, y = rmse, fill = method)) +
    geom_col(position = position_dodge(width = 0.72), width = 0.65, colour = "white") +
    scale_fill_manual(values = c("OLS"="#1B4965","Yeo-Johnson"="#E05C2A","Changepoint"="#2C6E8A")) +
    labs(title = "RMSE by method — smaller is better",
         subtitle = "Differences < 2% → linear regression is adequate",
         x = NULL, y = "RMSE", fill = NULL) +
    theme_minimal(base_size = 11, base_family = "serif") +
    theme(axis.text.x = element_text(angle = 35, hjust = 1),
          panel.grid.minor = element_blank(),
          plot.title    = element_text(face = "bold", size = 13),
          plot.subtitle = element_text(size = 10.5, colour = "#555"),
          legend.position = "top")
} else {
  htmltools::HTML("<div class='caja-discusion'>RMSE comparison not available.</div>")
}
```

### Detailed comparison table {data-width=440}

```{r}
if (exists("table_acceleration_dashboard") && nrow(table_acceleration_dashboard) > 0) {
  table_alt_condensed <- table_acceleration_dashboard %>%
    select(AREA, `Historical ± CI95.HAC (°C/century)`,
           `YJ recent (°C/century)`, `CP slope 2 (°C/century)`,
           `YJ improv. (%)`, `CP improv. (%)`,
           `YJ/hist. ratio`, `CP/hist. ratio`)
  DT::datatable(table_alt_condensed, rownames=FALSE, class="compact stripe",
    options=list(paging=FALSE, searching=FALSE, info=FALSE,
                 ordering=TRUE, autoWidth=TRUE, scrollX=TRUE))
} else if (has_yj) {
  table_yj <- acceleration %>%
    transmute(AREA,
      `Historical ± CI95.HAC (°C/century)` = fmt_century(slope_hist, ci95_hac),
      `YJ recent (°C/century)` = round(yj_recent*100, 2),
      `YJ/hist. ratio`         = round(yj_recent/slope_hist, 2))
  DT::datatable(table_yj, rownames=FALSE, class="compact stripe",
    options=list(paging=FALSE, searching=FALSE, info=FALSE,
                 ordering=TRUE, autoWidth=TRUE))
} else {
  htmltools::HTML("<div class='caja-discusion'>Table not available.</div>")
}
```


