---
title: "Temperature anomaly in the Canary Islands / World"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: readable
runtime: shiny
---

```{r global, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo    = FALSE,
  error   = FALSE
)
options(scipen = 999)
```

<style>
:root {
  --azul-oscuro: #1B4965;
  --azul-medio:  #2C6E8A;
  --naranja:     #E05C2A;
  --arena:       #8B7355;
  --fondo:       #f4f6f9;
  --blanco:      #ffffff;
  --gris-texto:  #555;
}

body {
  font-family: 'Segoe UI', system-ui, sans-serif;
  font-size: 14px;
  background-color: var(--fondo);
  color: #333;
}

/* ── Navbar ──────────────────────────────────────────────────────────── */
.navbar {
  background-color: var(--azul-oscuro) !important;
  border-bottom: 3px solid var(--naranja);
}
.navbar-brand {
  color: #fff !important;
  font-weight: 700 !important;
  font-size: 15px !important;
  letter-spacing: 0.02em;
}
.navbar-nav > li > a {
  color: #fff !important;
  font-weight: 600;
  letter-spacing: 0.03em;
}
.navbar-nav > li.active > a {
  background-color: var(--naranja) !important;
  color: #fff !important;
  border-radius: 4px;
}

/* ── Panels ──────────────────────────────────────────────────────────── */
.chart-wrapper, .chart-stage {
  background: var(--blanco);
  border-radius: 8px;
  box-shadow: 0 2px 8px rgba(0,0,0,0.07);
}
.chart-title {
  font-size: 12px;
  font-weight: 700;
  color: var(--azul-oscuro);
  padding: 6px 12px 3px;
  border-bottom: 2px solid var(--naranja);
  margin-bottom: 4px;
}

/* ── Sidebar ─────────────────────────────────────────────────────────── */
.section.sidebar {
  background: #e9eef3;
  border-right: 1px solid #cdd6df;
  padding: 14px 12px;
}
.section.sidebar h3 {
  color: var(--azul-oscuro);
  font-size: 13px;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: .06em;
  margin-bottom: 6px;
}

/* ── Value boxes ─────────────────────────────────────────────────────── */
.value-box {
  border-radius: 8px !important;
  box-shadow: 0 2px 8px rgba(0,0,0,0.10) !important;
}
.value-box .value   { font-size: 26px !important; font-weight: 700 !important; }
.value-box .caption { font-size: 12px !important; }

/* ── Interpretive boxes ──────────────────────────────────────────────── */
.caja-discusion {
  background: #fff8f5;
  border-left: 4px solid var(--naranja);
  padding: 10px 14px;
  margin: 8px 0;
  border-radius: 0 6px 6px 0;
  font-size: 13px;
  line-height: 1.6;
}
.caja-hallazgo {
  background: #eef6f8;
  border-left: 4px solid var(--azul-oscuro);
  padding: 10px 14px;
  margin: 8px 0;
  border-radius: 0 6px 6px 0;
  font-size: 13px;
  line-height: 1.6;
}
.caja-metodo {
  background: #f5f0eb;
  border-left: 4px solid var(--arena);
  padding: 10px 14px;
  margin: 8px 0;
  border-radius: 0 6px 6px 0;
  font-size: 13px;
  line-height: 1.6;
}
.caja-conclusion {
  background: #f0f4f8;
  border: 1px solid #c5d5e0;
  border-top: 3px solid var(--azul-oscuro);
  border-radius: 0 0 6px 6px;
  padding: 10px 14px;
  margin-top: 10px;
  font-size: 13px;
  line-height: 1.65;
}
.caja-conclusion .label {
  font-size: 10.5px;
  font-weight: 700;
  text-transform: uppercase;
  letter-spacing: .06em;
  color: var(--azul-oscuro);
  margin-bottom: 4px;
}

/* ── About compact redesign ───────────────────────────────────────── */
.about-onepage { padding: 0; }
.about-hero-wrap {
  background: linear-gradient(135deg, #f9fbfd 0%, #eef4f8 100%);
  border: 1px solid #d8e2e9;
  border-top: 4px solid var(--naranja);
  border-radius: 12px;
  padding: 10px 16px 12px 16px;
  box-shadow: 0 2px 6px rgba(0,0,0,0.04);
}
.about-topline {
  font-size: 10.5px; font-weight: 800; text-transform: uppercase;
  letter-spacing: .12em; color: var(--azul-medio); margin-bottom: 6px;
}
.about-main-title {
  font-size: 30px; line-height: 1.06; font-weight: 800;
  color: var(--azul-oscuro); margin: 0 0 6px 0;
}
.about-main-subtitle {
  font-size: 14px; line-height: 1.5; color: #516571;
  max-width: 1050px; margin-bottom: 6px;
}
.about-meta-line { font-size: 12.5px; color: #6b7b86; margin-bottom: 12px; }
.about-highlight-bar {
  background: #fff8f5; border-left: 4px solid var(--naranja);
  border-radius: 0 8px 8px 0; padding: 7px 11px; font-size: 12.6px;
  line-height: 1.55; color: #3f4b53; margin-bottom: 10px;
}
.about-content-grid {
  display: grid; grid-template-columns: 1.05fr 0.95fr;
  gap: 12px; align-items: stretch;
}
.about-copy-block {
  background: rgba(255,255,255,0.75); border: 1px solid #dde6ec;
  border-radius: 10px; padding: 14px 16px; display: flex;
  flex-direction: column; justify-content: space-between;
}
.about-copy-title { font-size: 17px; font-weight: 700; color: var(--azul-oscuro); margin-bottom: 8px; }
.about-copy-block p { font-size: 13px; line-height: 1.68; color: #3f4e57; margin: 0 0 8px 0; }
.about-copy-block p:last-child { margin-bottom: 0; }
.about-keyfacts {
  margin-top: 10px; padding-top: 8px; border-top: 1px solid #e4ebf0;
  font-size: 12.5px; line-height: 1.6; color: #5a6b76;
}
.about-reliability-box {
  margin-top: 10px; background: #f9fbfc; border: 1px solid #dbe6ed;
  border-radius: 10px; padding: 9px 11px;
}
.about-reliability-title {
  font-size: 11.5px; font-weight: 800; text-transform: uppercase;
  letter-spacing: .07em; color: var(--azul-oscuro); margin-bottom: 6px;
}
.about-reliability-intro { font-size: 12px; line-height: 1.5; color: #4e5f69; margin-bottom: 6px; }
.about-reliability-list { display: grid; grid-template-columns: 1fr; gap: 3px; }
.about-reliability-item { font-size: 11.8px; line-height: 1.42; color: #42515a; padding-left: 0; }
.about-reliability-item .tag { display: inline-block; min-width: 72px; font-weight: 700; color: var(--azul-oscuro); }
.about-map-card {
  background: rgba(255,255,255,0.82); border: 1px solid #dde6ec;
  border-radius: 10px; padding: 10px 10px 8px 10px; display: flex;
  flex-direction: column; min-height: 100%;
}
.about-map-title { font-size: 16px; font-weight: 700; color: var(--azul-oscuro); margin: 2px 2px 10px 2px; }
.about-map-caption { margin-top: 6px; font-size: 11.8px; line-height: 1.5; color: #6a7a86; }
@media (max-width: 1150px) {
  .about-content-grid { grid-template-columns: 1fr; }
  .about-main-title { font-size: 25px; }
}

/* ── KPI badge ───────────────────────────────────────────────────────── */
.kpi-badge { display: inline-block; padding: 2px 8px; border-radius: 12px; font-weight: 700; font-size: 12px; }
.kpi-ok  { background: #d4edda; color: #155724; }
.kpi-war { background: #fff3cd; color: #856404; }
.kpi-err { background: #f8d7da; color: #721c24; }

/* ── DT tables ───────────────────────────────────────────────────────── */
.dataTables_wrapper { font-size: 12.5px; }
table.dataTable thead th { background: var(--azul-oscuro); color: #fff; font-weight: 600; }

/* ── Secondary tabs ──────────────────────────────────────────────────── */
.nav-tabs > li.active > a,
.nav-tabs > li.active > a:hover {
  color: var(--azul-oscuro) !important;
  border-bottom: 3px solid var(--naranja) !important;
  font-weight: 700;
}
.nav-tabs > li > a { color: var(--gris-texto); }
.nav-tabs { margin-bottom: 6px !important; }

/* ── Select input ────────────────────────────────────────────────────── */
.selectize-input { border: 1px solid #b5c4cf !important; border-radius: 5px !important; font-size: 13px !important; }

/* ── Tabset fill ─────────────────────────────────────────────────────── */
.tabset-fill { height: 100%; }
.tabset-fill .tab-content { height: calc(100% - 52px); padding-top: 4px !important; }
.tabset-fill .tab-pane    { height: 100%; }
.tabset-fill .dataTables_wrapper,
.tabset-fill .dataTables_scroll,
.tabset-fill .dataTables_scrollBody { height: 100% !important; }
.tabset-fill .dataTables_scrollBody { min-height: 280px; }
.nota-bajo-tabset { margin-top: 8px; font-size: 11.5px; }

/* ── Alt models redesign ─────────────────────────────────────────────── */
.finding-box {
  background: #eef6f8; border: 1px solid #b8d4e0;
  border-top: 4px solid var(--azul-oscuro); border-radius: 0 0 8px 8px;
  padding: 12px 16px; font-size: 13px; line-height: 1.7; margin-top: 6px;
}
.finding-box .finding-title {
  font-size: 11px; font-weight: 700; text-transform: uppercase;
  letter-spacing: .07em; color: var(--azul-oscuro); margin-bottom: 6px;
}
.ratio-pill {
  display: inline-block; padding: 1px 8px; border-radius: 10px;
  font-weight: 700; font-size: 12px; margin: 1px 2px;
}
.ratio-accel  { background: #fde8df; color: #c0392b; }
.ratio-decel  { background: #dff0fb; color: #1B4965; }
.ratio-stable { background: #f0f0f0; color: #555; }

/* ── Trend metric cards ───────────────────────────────────────────── */
.trend-metrics-grid {
  display: grid; grid-template-columns: repeat(4, minmax(120px, 1fr));
  gap: 10px; margin-top: 10px;
}
.trend-metric-card {
  background: #f8fbfc; border: 1px solid #d7e2ea;
  border-top: 3px solid var(--azul-oscuro); border-radius: 8px;
  padding: 10px 12px; min-height: 78px;
}
.trend-metric-label {
  font-size: 11px; font-weight: 800; text-transform: uppercase;
  letter-spacing: .06em; color: var(--azul-oscuro); margin-bottom: 6px;
}
.trend-metric-value { font-size: 12.8px; line-height: 1.45; color: #30424d; font-weight: 600; }
.trend-metric-header { font-size: 15px; font-weight: 700; color: var(--azul-oscuro); margin-top: 10px; margin-bottom: 2px; }
@media (max-width: 1200px) { .trend-metrics-grid { grid-template-columns: repeat(2, minmax(120px, 1fr)); } }
@media (max-width: 700px)  { .trend-metrics-grid { grid-template-columns: 1fr; } }

/* ════════════════════════════════════════════════════════════════════
   SEASONALITY  – rediseño completo
   ════════════════════════════════════════════════════════════════════ */

.seas-shell {
  padding: 12px 4px 4px 4px;
  display: flex;
  flex-direction: column;
  gap: 12px;
}

/* Banner de contexto superior */
.seas-banner {
  background: linear-gradient(135deg, #f9fbfd 0%, #eef4f8 100%);
  border: 1px solid #d0dce6;
  border-left: 5px solid var(--naranja);
  border-radius: 8px;
  padding: 10px 16px;
  display: flex;
  align-items: flex-start;
  gap: 12px;
}
.seas-banner-icon { font-size: 22px; line-height: 1; padding-top: 2px; flex-shrink: 0; }
.seas-banner-text { font-size: 12.8px; line-height: 1.6; color: #3d4f5a; }
.seas-banner-text b { color: var(--azul-oscuro); }

/* Contenedor de los dos heatmaps lado a lado */
.seas-heatmap-grid {
  display: grid;
  grid-template-columns: 1fr 1fr;
  gap: 14px;
  align-items: start;
}

/* Cada tarjeta de heatmap */
.seas-hm-card {
  background: #fff;
  border: 1px solid #d8e2e9;
  border-radius: 10px;
  padding: 14px 14px 10px 14px;
  box-shadow: 0 2px 6px rgba(0,0,0,0.04);
  overflow: hidden;
}
.seas-hm-title {
  font-size: 13px; font-weight: 700; color: var(--azul-oscuro);
  margin-bottom: 2px;
}
.seas-hm-subtitle { font-size: 11.5px; color: #7a8b96; margin-bottom: 8px; }

/* Wrapper del plot con altura fija y scroll vertical si desborda */
.seas-hm-plot-wrap {
  width: 100%;
  overflow-x: auto;
  overflow-y: hidden;
}

/* Facet: toolbar + plot */
.seas-facet-shell {
  display: flex;
  flex-direction: column;
  gap: 10px;
}
.seas-facet-topbar {
  display: grid;
  grid-template-columns: 260px 1fr;
  gap: 12px;
  align-items: stretch;
}
.seas-facet-select-card {
  background: #fff;
  border: 1px solid #d8e2e9;
  border-top: 4px solid var(--arena);
  border-radius: 10px;
  padding: 12px 14px;
  box-shadow: 0 2px 6px rgba(0,0,0,0.04);
}
.seas-facet-select-title { font-size: 13px; font-weight: 700; color: var(--azul-oscuro); margin-bottom: 8px; }
.seas-facet-note {
  background: #eef6f8;
  border-left: 4px solid var(--azul-oscuro);
  border-radius: 0 8px 8px 0;
  padding: 10px 14px;
  font-size: 12.5px;
  line-height: 1.6;
  color: #3d4f5a;
}
.seas-facet-plot-card {
  background: #fff;
  border: 1px solid #d8e2e9;
  border-radius: 10px;
  padding: 12px 14px;
  box-shadow: 0 2px 6px rgba(0,0,0,0.04);
  overflow: auto;
}
.seas-facet-plot-title { font-size: 14px; font-weight: 700; color: var(--azul-oscuro); margin-bottom: 2px; }
.seas-facet-plot-legend { font-size: 11.5px; color: #7a8b96; margin-bottom: 8px; }

@media (max-width: 1100px) {
  .seas-heatmap-grid    { grid-template-columns: 1fr; }
  .seas-facet-topbar    { grid-template-columns: 1fr; }
}

/* ════════════════════════════════════════════════════════════════════
   ALTERNATIVE MODELS – rediseño completo
   ════════════════════════════════════════════════════════════════════ */

.alt-shell {
  padding: 12px 4px 4px 4px;
  display: flex;
  flex-direction: column;
  gap: 14px;
}

/* Tarjeta de resumen ejecutivo */
.alt-exec-card {
  background: linear-gradient(135deg, #f9fbfd 0%, #eef4f8 100%);
  border: 1px solid #d0dce6;
  border-top: 5px solid var(--azul-oscuro);
  border-radius: 10px;
  padding: 14px 18px 12px 18px;
  box-shadow: 0 2px 6px rgba(0,0,0,0.04);
}
.alt-exec-title {
  font-size: 16px; font-weight: 800; color: var(--azul-oscuro);
  margin-bottom: 6px; letter-spacing: -0.01em;
}
.alt-exec-lead {
  font-size: 13px; line-height: 1.65; color: #3d4f5a;
  max-width: 900px;
}

/* Tarjetas de hallazgo en grid */
.alt-findings-grid {
  display: grid;
  grid-template-columns: repeat(3, 1fr);
  gap: 12px;
}
.alt-finding-card {
  background: #fff;
  border: 1px solid #d8e2e9;
  border-radius: 10px;
  padding: 14px 16px;
  box-shadow: 0 2px 5px rgba(0,0,0,0.04);
  position: relative;
}
.alt-finding-card::before {
  content: '';
  position: absolute;
  top: 0; left: 0; right: 0;
  height: 4px;
  border-radius: 10px 10px 0 0;
}
.alt-finding-card.green::before  { background: #27ae60; }
.alt-finding-card.orange::before { background: var(--naranja); }
.alt-finding-card.blue::before   { background: var(--azul-oscuro); }

.alt-finding-icon  { font-size: 20px; margin-bottom: 6px; }
.alt-finding-title { font-size: 12px; font-weight: 800; text-transform: uppercase; letter-spacing: .06em; color: var(--azul-oscuro); margin-bottom: 6px; }
.alt-finding-body  { font-size: 12.5px; line-height: 1.6; color: #3d4f5a; }
.alt-finding-body b { color: var(--azul-oscuro); }
.alt-finding-tag {
  display: inline-block; margin-top: 8px; padding: 2px 8px;
  border-radius: 10px; font-size: 11.5px; font-weight: 700;
}
.tag-green  { background: #d4edda; color: #1a5e32; }
.tag-orange { background: #fde8df; color: #c0392b; }
.tag-blue   { background: #dff0fb; color: #1B4965; }

/* Grid de plots + tabla */
.alt-plots-grid {
  display: grid;
  grid-template-columns: 1.2fr 0.8fr;
  gap: 14px;
  align-items: start;
}
.alt-plot-card {
  background: #fff; border: 1px solid #d8e2e9; border-radius: 10px;
  padding: 12px 14px; box-shadow: 0 2px 5px rgba(0,0,0,0.04);
}
.alt-plot-title    { font-size: 13px; font-weight: 700; color: var(--azul-oscuro); margin-bottom: 2px; }
.alt-plot-subtitle { font-size: 11.5px; color: #7a8b96; margin-bottom: 8px; }

.alt-table-card {
  background: #fff; border: 1px solid #d8e2e9; border-radius: 10px;
  padding: 12px 14px; box-shadow: 0 2px 5px rgba(0,0,0,0.04);
}
.alt-table-title    { font-size: 13px; font-weight: 700; color: var(--azul-oscuro); margin-bottom: 2px; }
.alt-table-subtitle { font-size: 11.5px; color: #7a8b96; margin-bottom: 8px; }

.alt-ratio-table {
  width: 100%; border-collapse: collapse; font-size: 12px;
}
.alt-ratio-table thead th {
  background: var(--azul-oscuro); color: #fff;
  padding: 6px 8px; text-align: left; font-weight: 600;
}
.alt-ratio-table tbody td {
  padding: 5px 8px; border-bottom: 1px solid #e8edf2;
}
.alt-ratio-table tbody tr:hover { background: #f5f8fb; }

.alt-rmse-note {
  background: #f0f4f8; border-left: 3px solid var(--arena);
  border-radius: 0 6px 6px 0; padding: 8px 12px;
  font-size: 12px; line-height: 1.55; color: #4d5f6a; margin-top: 10px;
}

@media (max-width: 1100px) {
  .alt-findings-grid { grid-template-columns: 1fr 1fr; }
  .alt-plots-grid    { grid-template-columns: 1fr; }
}
@media (max-width: 700px) {
  .alt-findings-grid { grid-template-columns: 1fr; }
}
</style>

```{r libs}
library(shiny)
library(flexdashboard)
library(tidyverse)
library(lubridate)
library(tsibble)
library(fpp3)
library(purrr)
library(zoo)
library(ggplot2)
library(plotly)
library(leaflet)
library(broom)
library(lmtest)
library(sandwich)
library(trend)
library(nortest)
library(MASS)
library(tseries)
library(knitr)
library(DT)

source("../scripts/utilities.R")
```

```{r ns_fix}
select      <- dplyr::select
mutate      <- dplyr::mutate
filter      <- dplyr::filter
group_by    <- dplyr::group_by
summarise   <- dplyr::summarise
selectInput <- shiny::selectInput
```

```{r params}
START_DATE     <- as.Date("1970-01-01")
END_DATE       <- as.Date("2024-12-31")
N_RECENT_YEARS <- 10
CACHE_VERSION  <- "v2"

area_order <- c(
  "GLOBAL", "NORTHERN HEMISPHERE", "SOUTHERN HEMISPHERE",
  "CANARY ISLANDS", "GRAN CANARIA/GANDO", "IZAÑA",
  "S.C. DE TENERIFE", "TENERIFE/LOS RODEOS"
)

MONTH_NAMES     <- c("January","February","March","April","May","June",
                     "July","August","September","October","November","December")
month_levels_en <- tolower(MONTH_NAMES)

# ── AEMET station coordinates ────────────────────────────────────────────────
stations_df <- tibble(
  name    = c("IZAÑA", "TENERIFE/LOS RODEOS", "S.C. DE TENERIFE", "GRAN CANARIA/GANDO"),
  label   = c("Izaña\n(2,373 m)", "Tenerife/Los Rodeos\n(632 m)",
              "S.C. de Tenerife\n(36 m)", "Gran Canaria/Gando\n(24 m)"),
  lat     = c(28.309,  28.478,  28.463,  27.926),
  lon     = c(-16.499, -16.341, -16.255, -15.385),
  alt_m   = c(2373,    632,     36,      24),
  island  = c("Tenerife", "Tenerife", "Tenerife", "Gran Canaria"),
  since   = c("1916", "1944", "1920", "1950")
)
```

```{r helpers}
fmt_century <- function(slope, ic, digits = 2) {
  paste0(
    formatC(round(slope * 100, digits), format = "f", digits = digits),
    " ± ",
    formatC(round(ic * 100, digits), format = "f", digits = digits)
  )
}

ensure_yj_cols <- function(df) {
  yj_cols <- c("RMSE.YJ","SW.YJ","AD.YJ","BP.YJ","lambda.YJ","trend.YJ")
  for (col in yj_cols) if (!col %in% names(df)) df[[col]] <- NA_real_
  df
}

calc_reliability <- function(r2, ic, slope, ad, bp, mk, lb) {
  flags <- c(
    if (!is.na(r2) && r2 < 0.10)                        "⛔R²<0.10",
    if (!is.na(ic) && !is.na(slope) && ic > abs(slope)) "⚠️CI>slope",
    if (!is.na(ad) && ad < 0.05)                        "⚠️AD",
    if (!is.na(bp) && bp < 0.05)                        "⚠️BP",
    if (!is.na(mk) && mk > 0.05)                        "⚠️MK n.s.",
    if (!is.na(lb) && lb < 0.05)                        "⚠️LB"
  )
  if (length(flags) == 0) "✅ OK" else paste(flags, collapse = " ")
}

calc_reliability_num <- function(r2, ic, slope, ad, bp, mk, lb) {
  if (!is.na(r2) && r2 < 0.10) return(2L)
  nfaults <- 0L
  if (!is.na(ic) && !is.na(slope) && ic > abs(slope)) nfaults <- nfaults + 1L
  if (!is.na(ad) && ad < 0.05) nfaults <- nfaults + 1L
  if (!is.na(bp) && bp < 0.05) nfaults <- nfaults + 1L
  if (!is.na(mk) && mk > 0.05) nfaults <- nfaults + 1L
  if (!is.na(lb) && lb < 0.05) nfaults <- nfaults + 1L
  nfaults
}

calc_reliability_label <- function(r2, ic, slope, ad, bp, mk, lb) {
  if (!is.na(r2) && r2 < 0.10)                        return("⛔R²")
  if (!is.na(ic) && !is.na(slope) && ic > abs(slope)) return("⚠️CI")
  if (!is.na(ad) && ad < 0.05)                        return("⚠️AD")
  if (!is.na(bp) && bp < 0.05)                        return("⚠️BP")
  if (!is.na(mk) && mk > 0.05)                        return("⚠️MK")
  if (!is.na(lb) && lb < 0.05)                        return("⚠️LB")
  "✅OK"
}

# ── Plot functions ────────────────────────────────────────────────────────────

plot_heatmap_slope <- function(df) {
  df %>%
    ggplot(aes(x = period, y = AREA, fill = slope_century)) +
    geom_tile(colour = "white", linewidth = 0.7) +
    geom_text(aes(label = sprintf("%.2f", slope_century)), size = 3.0, colour = "black") +
    scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "°C/century") +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 11, base_family = "sans") +
    theme(
      axis.text.x  = element_text(angle = 40, hjust = 1, size = 9),
      axis.text.y  = element_text(size = 9),
      panel.grid   = element_blank(),
      legend.position = "right",
      plot.margin  = margin(4, 4, 4, 4)
    )
}

plot_heatmap_reliability <- function(df) {
  df %>%
    ggplot(aes(x = month_name, y = AREA, fill = reliability_num, label = reliability_label)) +
    geom_tile(colour = "white", linewidth = 0.7) +
    geom_text(size = 2.9, colour = "black") +
    scale_fill_gradient(low = "chartreuse1", high = "firebrick1") +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 11, base_family = "sans") +
    theme(
      axis.text.x  = element_text(angle = 40, hjust = 1, size = 9),
      axis.text.y  = element_text(size = 9),
      panel.grid   = element_blank(),
      legend.position = "none",
      plot.margin  = margin(4, 4, 4, 4)
    )
}

plot_facet_monthly <- function(data_ts, area_sel) {
  bands_area <- data_ts %>%
    as_tibble() %>%
    filter(!is.na(value), AREA == area_sel) %>%
    mutate(
      date       = as.Date(date),
      month_name = MONTH_NAMES[month(date)],
      month_name = factor(month_name, levels = MONTH_NAMES)
    ) %>%
    group_by(month_name) %>%
    group_modify(~ {
      mod      <- lm(value ~ date, data = .x)
      pred_ols <- predict(mod, interval = "confidence", level = 0.95)
      V_hac    <- sandwich::vcovHAC(mod)
      X        <- model.matrix(mod)
      se_hac   <- sqrt(rowSums((X %*% V_hac) * X))
      tc       <- qt(0.975, df = mod$df.residual)
      tibble(
        date   = .x$date,
        value  = .x$value,
        fit    = pred_ols[, "fit"],
        ols_lo = pred_ols[, "lwr"],
        ols_hi = pred_ols[, "upr"],
        hac_lo = pred_ols[, "fit"] - tc * se_hac,
        hac_hi = pred_ols[, "fit"] + tc * se_hac
      )
    }) %>%
    ungroup()

  if (nrow(bands_area) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5, label = paste("No data for:", area_sel),
                 size = 5, colour = "#1B4965") +
        theme_void()
    )
  }

  ggplot(bands_area, aes(x = date)) +
    geom_ribbon(aes(ymin = hac_lo, ymax = hac_hi), alpha = 0.18, fill = "#E05C2A") +
    geom_line(aes(y = value), linewidth = 0.35, alpha = 0.50, colour = "grey55") +
    geom_line(aes(y = fit), colour = "#1B4965", linewidth = 0.9) +
    facet_wrap(~ month_name, ncol = 4, scales = "free_y") +
    labs(x = NULL, y = "Anomaly (°C)") +
    theme_minimal(base_size = 10, base_family = "sans") +
    theme(
      strip.text       = element_text(face = "bold", size = 9, colour = "#1B4965"),
      strip.background = element_rect(fill = "#eef4f8", colour = NA),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "#e8edf2"),
      axis.text.x      = element_text(size = 7),
      axis.text.y      = element_text(size = 7),
      axis.title.y     = element_text(size = 9),
      panel.spacing    = unit(0.6, "lines"),
      plot.margin      = margin(4, 6, 4, 8)
    )
}
```

```{r load_data}
all_data <- readRDS("../data/temperature_anomaly_2025.rds") %>%
  as_tibble() %>%
  filter(as.Date(date) >= START_DATE, as.Date(date) <= END_DATE)

levels(all_data$AREA)[levels(all_data$AREA) == "LAS PALMAS DE GRAN CANARIA/GANDO"] <- "GRAN CANARIA/GANDO"
levels(all_data$AREA)[levels(all_data$AREA) == "SANTA CRUZ DE TENERIFE"]           <- "S.C. DE TENERIFE"
levels(all_data$AREA)[levels(all_data$AREA) == "NH"]      <- "NORTHERN HEMISPHERE"
levels(all_data$AREA)[levels(all_data$AREA) == "SH"]      <- "SOUTHERN HEMISPHERE"
levels(all_data$AREA)[levels(all_data$AREA) == "CANARIAS"] <- "CANARY ISLANDS"

all_data <- all_data %>%
  mutate(AREA = factor(AREA, levels = area_order)) %>%
  as_tsibble(index = date, key = AREA) %>%
  fill_gaps()

all_data_num <- all_data %>%
  as_tibble() %>%
  mutate(time = as.numeric(date), AREA = factor(AREA, levels = area_order))

all_data_annual <- all_data %>%
  as_tibble() %>%
  mutate(year = year(date), AREA = factor(AREA, levels = area_order)) %>%
  group_by(AREA, year) %>%
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(value))

all_data_monthly <- all_data %>%
  as_tibble() %>%
  mutate(year = year(date), month = month(date), AREA = factor(AREA, levels = area_order)) %>%
  filter(!is.na(value))
```

```{r load_precomputed, include=FALSE}
results <- readRDS("../data/dashboard_results.rds")

results_lm               <- results$results_lm
results_annual           <- results$results_annual
results_monthly          <- results$results_monthly
results_monthly2         <- results$results_monthly2
table_annual_dashboard   <- results$table_annual_dashboard
heatmap_slope_data       <- results$heatmap_slope_data
heatmap_reliability_data <- results$heatmap_reliability_data
diagnostic_summary       <- results$diagnostic_summary
bands_monthly            <- results$bands_monthly
bands_annual             <- results$bands_annual
acceleration             <- results$acceleration
table_acceleration_dashboard <- results$table_acceleration_dashboard
best_cp                  <- results$best_cp
kpi_canary_annual        <- results$kpi_canary_annual
kpi_global_annual        <- results$kpi_global_annual
kpi_nh_annual            <- results$kpi_nh_annual
kpi_sh_annual            <- results$kpi_sh_annual
kpi_n_areas_sig          <- results$kpi_n_areas_sig
kpi_changepoint          <- results$kpi_changepoint
kpi_max_improvement      <- results$kpi_max_improvement
```



<!-- =====================================================================
     TAB 0 – ABOUT
     ===================================================================== -->

About
=====================================

Row {data-height=640}
-------------------------------------

### {.no-title}

```{r}
htmltools::tagList(

  htmltools::HTML('
  <div class="about-onepage">
    <div class="about-hero-wrap">

      <div class="about-main-title">
        Temperature Anomaly Dashboard · Canary Islands & Global Context (1970–2024)
      </div>

      <div class="about-main-subtitle">
        Long-term warming trends in the Canary Islands are analysed in a global context
        using monthly temperature anomaly series and robust statistical diagnostics.
      </div>

      <div class="about-highlight-bar">
        <b>Main finding.</b> All analysed areas show significant warming, with the Canary Islands
        regional trend broadly comparable to the Northern Hemisphere and clear seasonal differences across stations.
      </div>

      <div class="about-content-grid">
        <div class="about-copy-block">
          <div>
            <div class="about-copy-title">Project overview</div>
            <p>
              This dashboard analyses temperature anomaly trends for the global mean,
              both hemispheres, four long-record AEMET stations, and a regional Canary Islands mean,
              combining monthly, annual, and month-specific perspectives to capture both
              <b>long-term warming intensity</b> and <b>seasonal structure</b> over the period <b>1970–2024</b>.
            </p>
            <p>
              A temperature anomaly is the <b>difference between the observed temperature and the reference-period mean</b> for the same month.
            </p>
            <p>
              The analysis uses data from <b>HadCRUT5</b> and <b>AEMET</b> and is supported by
              <b>HAC confidence intervals</b>, <b>Sen\'s slope</b>, <b>Mann-Kendall testing</b>,
              and standard residual diagnostics, while alternative models are included only
              to assess <b>recent-trend changes</b> and departures from linearity.
            </p>
            <div class="about-reliability-box">
              <div class="about-reliability-title">Reliability indicators</div>
              <div class="about-reliability-intro">
                Each model receives an automatic diagnostic label. When no warning is triggered,
                the model is classified as <b>✅ OK</b>.
              </div>
              <div class="about-reliability-list">
                <div class="about-reliability-item"><span class="tag">⛔ R²&lt;0.10</span>trend signal dominated by noise.</div>
                <div class="about-reliability-item"><span class="tag">⚠️ CI&gt;slope</span>trend not clearly distinguishable from zero.</div>
                <div class="about-reliability-item"><span class="tag">⚠️ AD</span>residual normality rejected.</div>
                <div class="about-reliability-item"><span class="tag">⚠️ BP</span>heteroscedasticity detected.</div>
                <div class="about-reliability-item"><span class="tag">⚠️ LB</span>residual autocorrelation detected.</div>
                <div class="about-reliability-item"><span class="tag">⚠️ MK n.s.</span>Mann-Kendall does not detect a significant monotonic trend.</div>
              </div>
            </div>
          </div>
        </div>

        <div class="about-map-card">
          <div class="about-map-title">AEMET stations used in the study</div>
  '),

  leaflet(stations_df, height = 400) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%
    setView(lng = -15.9, lat = 28.2, zoom = 7) %>%
    addCircleMarkers(
      lng = ~lon, lat = ~lat,
      radius = ~pmax(5, sqrt(alt_m) * 0.22),
      color = "#1B4965", fillColor = "#E05C2A", fillOpacity = 0.72, weight = 1.4,
      popup = ~paste0("<b>", name, "</b><br>Island: ", island, "<br>Altitude: ",
                      format(alt_m, big.mark=","), " m a.s.l.<br>Records since: ", since),
      label = ~name
    ) %>%
    addLegend(position = "bottomright", colors = "#E05C2A",
              labels = "AEMET station", title = "Study network", opacity = 0.85),

  htmltools::HTML('
          <div class="about-map-caption">
            Four long-term AEMET stations selected to represent contrasting coastal and high-altitude environments.
          </div>
        </div>
      </div>
    </div>
  </div>
  ')
)
```


<!-- =====================================================================
     TAB 1 – EXECUTIVE SUMMARY
     ===================================================================== -->

Executive Summary
=====================================

Row {data-height=130}
-------------------------------------

### 🌴 Canary Islands

```{r}
valueBox(value   = paste0(round(kpi_canary_annual * 100, 2), " °C/century"),
         caption = "Annual slope · Canary Islands",
         icon    = "fa-thermometer-half", color = "#1B4965")
```

### 🌍 Global

```{r}
valueBox(value   = paste0(round(kpi_global_annual * 100, 2), " °C/century"),
         caption = "Annual slope · Global",
         icon    = "fa-globe", color = "#2C6E8A")
```

### 🧭 Northern Hemisphere

```{r}
valueBox(value   = paste0(round(kpi_nh_annual * 100, 2), " °C/century"),
         caption = "Annual slope · Northern Hemisphere",
         icon    = "fa-line-chart", color = "#8B7355")
```

### 🧭 Southern Hemisphere

```{r}
valueBox(
  value   = paste0(round(kpi_sh_annual * 100, 2), " °C/century"),
  caption = "Annual slope · Southern Hemisphere",
  icon    = "fa-arrow-down", color = "#E05C2A"
)
```

Row
-------------------------------------

### Annual slope comparison by area {data-width=600}

```{r}
bars_data <- results_annual %>%
  mutate(AREA = factor(AREA, levels = rev(area_order)),
         slope_century = lm.slope * 100, ci_century = CI95.HAC * 100,
         tooltip_txt = paste0(
           "<b>", AREA, "</b><br>",
           "OLS slope: <b>", round(lm.slope*100,2), " °C/century</b><br>",
           "HAC CI95: ± ", round(CI95.HAC*100,2), "<br>",
           "R²: ", round(R2,3), "<br>",
           "Reliability: ", reliability))

plot_ly(data=bars_data, x=~slope_century, y=~AREA, type="bar", orientation="h",
        error_x=list(type="data", array=~ci_century, color="#E05C2A", thickness=1.5),
        marker=list(color="#1B4965"), hovertext=~tooltip_txt, hoverinfo="text") %>%
  layout(title  = list(text="Annual slope (°C/century) · OLS with HAC CI95", font=list(size=14)),
         xaxis  = list(title="°C/century", zeroline=TRUE, zerolinecolor="#aaa"),
         yaxis  = list(title=""),
         margin = list(l=180, r=20, t=50, b=40),
         plot_bgcolor="rgba(0,0,0,0)", paper_bgcolor="rgba(0,0,0,0)") %>%
  config(displayModeBar=FALSE)
```

### Quick diagnostics {data-width=400}

```{r}
diagnostic_summary_t <- diagnostic_summary %>%
  mutate(scale = as.character(scale)) %>%
  tibble::column_to_rownames("scale") %>%
  t() %>%
  as.data.frame(check.names = FALSE) %>%
  tibble::rownames_to_column("Metric")

div(class = "tabset-fill", tagList(
  tabsetPanel(
    tabPanel("📋 Summary table",
      tagList(
        DT::datatable(
          table_annual_dashboard,
          rownames = FALSE,
          class = "compact stripe",
          options = list(paging=FALSE, searching=FALSE, info=FALSE,
                         ordering=TRUE, autoWidth=TRUE, scrollX=FALSE)
        ),
        tags$div(
          class = "caja-metodo nota-bajo-tabset",
          HTML("<b>HAC CI:</b> corrected for autocorrelation and heteroscedasticity (Newey-West).
               <b>Sen slope:</b> robust Theil-Sen estimator, resistant to outliers.
               Full indicator guide available in the <b>About</b> tab.")
        )
      )
    ),
    tabPanel("🔬 Diagnostics",
      DT::datatable(
        diagnostic_summary_t,
        rownames = FALSE,
        class = "compact stripe",
        options = list(paging=FALSE, searching=FALSE, info=FALSE,
                       ordering=FALSE, autoWidth=TRUE, scrollX=TRUE)
      ) %>% DT::formatStyle("Metric", fontWeight = "bold")
    )
  )
))
```


<!-- =====================================================================
     TAB 2 – TRENDS
     ===================================================================== -->

Trends
=====================================

Sidebar {.sidebar}
-------------------------------------

### Area selection

```{r}
selectInput(inputId = "area_trend", label = "Select area:",
            choices = area_order, selected = "CANARY ISLANDS")

renderUI({
  req(input$area_trend)
  row_m <- results_lm    %>% filter(AREA == input$area_trend) %>% slice(1)
  row_a <- results_annual %>% filter(AREA == input$area_trend) %>% slice(1)
  dif <- round((row_m$lm.slope - row_a$lm.slope) * 100, 2)
  dir <- if (dif > 0) "higher" else "lower"
  htmltools::HTML(paste0(
    "<div class='caja-conclusion' style='margin-top:8px'>",
    "<div class='label'>🔎 Summary</div>",
    "Monthly slope is <b>", dir, "</b> than annual by <b>", abs(dif), " °C/century</b>. ",
    "Notable differences suggest seasonality influences the overall estimate.",
    "</div>",
    "<div class='caja-metodo' style='margin-top:8px; font-size:11.5px'>",
    "<b>HAC CI:</b> corrects autocorrelation/heteroscedasticity.<br>",
    "<b>Sen:</b> robust estimator (Theil-Sen).",
    "</div>"
  ))
})
```

Row
-------------------------------------

### {.no-title}

```{r}
tabsetPanel(
  tabPanel("📅 Monthly series",
    renderUI({
      req(input$area_trend)
      row_m <- results_lm %>% filter(AREA == input$area_trend) %>% slice(1)
      badge_class <- function(rel) {
        if (grepl("⛔", rel)) "kpi-err" else if (grepl("⚠️", rel)) "kpi-war" else "kpi-ok"
      }
      htmltools::HTML(paste0(
        "<div class='caja-hallazgo' style='margin-bottom:10px'>",
        "<b>R²:</b> ", round(row_m$R2, 3),
        " &nbsp;|&nbsp; <b>OLS HAC:</b> ", fmt_century(row_m$lm.slope, row_m$CI95.HAC), " °C/century",
        " &nbsp;|&nbsp; <b>Sen:</b> ", fmt_century(row_m$sen.slope, row_m$sen.CI95), " °C/century",
        " &nbsp;|&nbsp; <b>Status:</b> <span class='kpi-badge ", badge_class(row_m$reliability), "'>",
        row_m$reliability, "</span></div>"
      ))
    }),
    renderPlotly({
      req(input$area_trend)
      d <- bands_monthly %>% filter(AREA == input$area_trend)
      plot_ly(d, x = ~date) %>%
        add_ribbons(ymin=~hac_lo, ymax=~hac_hi, fillcolor="rgba(224,92,42,0.15)",
                    line=list(color="transparent"), name="HAC CI95") %>%
        add_lines(y=~value, line=list(color="rgba(120,120,120,0.6)", width=0.8), name="Observed anomaly") %>%
        add_lines(y=~fit,   line=list(color="#1B4965", width=2.2), name="OLS trend") %>%
        layout(title=list(text=paste("Monthly series:", input$area_trend), font=list(size=14)),
               xaxis=list(title=""), yaxis=list(title="Anomaly (°C)"),
               legend=list(orientation="h", y=-0.18), hovermode="x unified",
               plot_bgcolor="rgba(0,0,0,0)", paper_bgcolor="rgba(0,0,0,0)") %>%
        config(displayModeBar=FALSE)
    })
  ),
  tabPanel("📆 Annual mean",
    renderUI({
      req(input$area_trend)
      row_a <- results_annual %>% filter(AREA == input$area_trend) %>% slice(1)
      badge_class <- function(rel) {
        if (grepl("⛔", rel)) "kpi-err" else if (grepl("⚠️", rel)) "kpi-war" else "kpi-ok"
      }
      htmltools::HTML(paste0(
        "<div class='caja-hallazgo' style='margin-bottom:10px'>",
        "<b>R²:</b> ", round(row_a$R2, 3),
        " &nbsp;|&nbsp; <b>OLS HAC:</b> ", fmt_century(row_a$lm.slope, row_a$CI95.HAC), " °C/century",
        " &nbsp;|&nbsp; <b>Sen:</b> ", fmt_century(row_a$sen.slope, row_a$sen.CI95), " °C/century",
        " &nbsp;|&nbsp; <b>Status:</b> <span class='kpi-badge ", badge_class(row_a$reliability), "'>",
        row_a$reliability, "</span></div>"
      ))
    }),
    renderPlotly({
      req(input$area_trend)
      d <- bands_annual %>% filter(AREA == input$area_trend) %>% mutate(date=as.Date(paste0(year,"-01-01")))
      plot_ly(d, x=~date) %>%
        add_ribbons(ymin=~hac_lo, ymax=~hac_hi, fillcolor="rgba(224,92,42,0.15)",
                    line=list(color="transparent"), name="HAC CI95") %>%
        add_lines(y=~value, line=list(color="rgba(100,100,100,0.65)", width=1), name="Observed annual mean") %>%
        add_markers(y=~value, marker=list(color="#8B7355", size=5), name="", showlegend=FALSE) %>%
        add_lines(y=~fit, line=list(color="#1B4965", width=2.5), name="OLS trend") %>%
        layout(title=list(text=paste("Annual mean:", input$area_trend), font=list(size=14)),
               xaxis=list(title=""), yaxis=list(title="Annual mean anomaly (°C)"),
               legend=list(orientation="h", y=-0.18), hovermode="x unified",
               plot_bgcolor="rgba(0,0,0,0)", paper_bgcolor="rgba(0,0,0,0)") %>%
        config(displayModeBar=FALSE)
    })
  )
)
```


<!-- =====================================================================
     TAB 3 – SEASONALITY  (rediseñado)
     ===================================================================== -->

Seasonality
=====================================

```{r}
output$heatmap_slope_plot <- renderPlot({
  plot_heatmap_slope(heatmap_slope_data)
}, res = 96)

output$heatmap_reliability_plot <- renderPlot({
  plot_heatmap_reliability(heatmap_reliability_data)
}, res = 96)

output$facet_monthly_plot <- renderPlot({
  req(input$area_month)
  plot_facet_monthly(all_data, input$area_month)
}, res = 96)

output$facet_title <- renderText({
  req(input$area_month)
  paste("Monthly anomaly evolution —", input$area_month)
})
```

Row
-------------------------------------

### {.no-title}

```{r}
tabsetPanel(

  # ── Heatmaps ────────────────────────────────────────────────────────
  tabPanel("🔥 Heatmaps",
    div(class = "seas-shell",

      # Banner contextual
      div(class = "seas-banner",
        div(class = "seas-banner-icon", "🗓️"),
        div(class = "seas-banner-text",
          HTML("<b>Does warming depend on the time of year?</b> Yes — and it varies by location.
               The left map shows how fast each area–month combination is warming (°C per century).
               The right map flags months where the statistical fit is less reliable.
               <b>Key takeaway:</b> Canarian stations warm faster in spring–summer; the Northern Hemisphere
               shows its strongest signal in winter. Reliability issues are concentrated in
               Izaña and Los Rodeos during winter months.")
        )
      ),

      # Grid de dos heatmaps
      div(class = "seas-heatmap-grid",

        div(class = "seas-hm-card",
          div(class = "seas-hm-title", "Warming intensity"),
          div(class = "seas-hm-subtitle",
              "OLS slope (°C/century) per area and calendar month — darker orange = faster warming"),
          div(class = "seas-hm-plot-wrap",
            plotOutput("heatmap_slope_plot", height = "420px", width = "100%")
          )
        ),

        div(class = "seas-hm-card",
          div(class = "seas-hm-title", "Statistical reliability"),
          div(class = "seas-hm-subtitle",
              "Green = model OK · Red = one or more diagnostic warnings — check the About tab for codes"),
          div(class = "seas-hm-plot-wrap",
            plotOutput("heatmap_reliability_plot", height = "420px", width = "100%")
          )
        )
      )
    )
  ),

  # ── Monthly facets ───────────────────────────────────────────────────
  tabPanel("📅 Monthly facets",
    div(class = "seas-shell",

      # Toolbar: selector + nota
      div(class = "seas-facet-topbar",

        div(class = "seas-facet-select-card",
          div(class = "seas-facet-select-title", "Select area"),
          selectInput(inputId="area_month", label=NULL, choices=area_order,
                      selected="CANARY ISLANDS", width="100%")
        ),

        div(class = "seas-facet-note",
          HTML("<b>What am I looking at?</b> Each panel shows one calendar month for the selected area.
               The <b style='color:#1B4965'>blue line</b> is the long-term warming trend (OLS).
               The <b style='color:#E05C2A'>orange band</b> is the 95% HAC confidence interval
               (corrected for autocorrelation). The grey line is the raw observed anomaly.
               Steeper blue lines → faster warming in that month.")
        )
      ),

      # Plot card — altura adaptativa con scroll
      div(class = "seas-facet-plot-card",
        div(class = "seas-facet-plot-title", textOutput("facet_title", inline = TRUE)),
        div(class = "seas-facet-plot-legend",
            "Grey: observed · Blue: OLS trend · Orange band: 95% HAC confidence interval"),
        div(style = "overflow-y: auto;",
          plotOutput("facet_monthly_plot", height = "680px", width = "100%")
        )
      )
    )
  )
)
```


<!-- =====================================================================
     TAB 4 – ALTERNATIVE MODELS  (rediseñado)
     ===================================================================== -->

Alternative Models
=====================================

```{r alt_flags, include=FALSE}
has_yj <- nrow(acceleration) > 0 &&
  all(c("AREA","slope_hist","yj_recent","rmse_ols","rmse_yj") %in% names(acceleration))
has_cp <- nrow(acceleration) > 0 &&
  all(c("slope2_cp","rmse_cp","changepoint") %in% names(acceleration))
```

Row
-------------------------------------

### {.no-title}

```{r}
if (has_yj) {

  # ── Pre-compute ratio data ───────────────────────────────────────────────
  ratio_data <- acceleration %>%
    mutate(
      AREA     = factor(AREA, levels = area_order),
      ratio_yj = round(yj_recent / slope_hist, 2),
      ratio_cp = if (has_cp) round(slope2_cp * 12 / slope_hist, 2) else NA_real_
    ) %>%
    arrange(AREA)

  pill <- function(r) {
    if (is.na(r)) return("<span style='color:#aaa'>—</span>")
    cls <- if (r > 1.05) "ratio-accel" else if (r < 0.95) "ratio-decel" else "ratio-stable"
    paste0("<span class='ratio-pill ", cls, "'>", r, "</span>")
  }

  ratio_rows <- ratio_data %>%
    pmap_chr(function(AREA, ratio_yj, ratio_cp, ...) {
      paste0("<tr><td>", AREA, "</td>",
             "<td style='text-align:center'>", pill(ratio_yj), "</td>",
             if (has_cp) paste0("<td style='text-align:center'>", pill(ratio_cp), "</td>") else "",
             "</tr>")
    }) %>%
    paste(collapse = "")

  header_cp <- if (has_cp) "<th>CP / hist.</th>" else ""

  htmltools::HTML(paste0('
  <div class="alt-shell">

    <!-- ── Resumen ejecutivo ──────────────────────────────────────── -->
    <div class="alt-exec-card">
      <div class="alt-exec-title">What are alternative models and why do they matter?</div>
      <div class="alt-exec-lead">
        The main analysis assumes warming is <em>constant and linear</em> over time. Here we test two
        alternatives that allow the trend to <strong>accelerate, slow down, or break</strong>:
        <strong>Yeo-Johnson</strong> (captures a gradual curve) and <strong>Changepoint / segmented regression</strong>
        (detects an abrupt break point in the trend). If these models do not improve the fit by more than 2%,
        the linear model is adequate — which is what we find for all areas.
      </div>
    </div>

    <!-- ── Tres hallazgos clave ───────────────────────────────────── -->
    <div class="alt-findings-grid">

      <div class="alt-finding-card green">
        <div class="alt-finding-icon">✅</div>
        <div class="alt-finding-title">Linear model is sufficient</div>
        <div class="alt-finding-body">
          Neither alternative model improves RMSE by more than <b>2%</b> in any area.
          This means the long-run warming trend is well captured by a straight line — no need
          for more complex models to describe the overall pattern.
        </div>
        <span class="alt-finding-tag tag-green">Max. RMSE improvement: ',
        round(kpi_max_improvement, 2), '%</span>
      </div>

      <div class="alt-finding-card orange">
        <div class="alt-finding-icon">🌍</div>
        <div class="alt-finding-title">Global warming is accelerating</div>
        <div class="alt-finding-body">
          Both methods agree: the recent warming rate for the globe and Northern Hemisphere is
          <b>~25–28% higher</b> than the historical average.
          The breakpoint occurred around <b>', kpi_changepoint, '</b>.
          In plain terms: warming is speeding up at planetary scale.
        </div>
        <span class="alt-finding-tag tag-orange">YJ ratio ≈ 1.27–1.28 · CP ratio ≈ 1.07–1.08</span>
      </div>

      <div class="alt-finding-card blue">
        <div class="alt-finding-icon">🏝️</div>
        <div class="alt-finding-title">Canary Islands: possible slowdown</div>
        <div class="alt-finding-body">
          Canarian stations show <b>ratios below 1</b> in both methods (0.58–0.96),
          suggesting the recent warming rate is somewhat <em>lower</em> than the historical average.
          The signal is weak but consistent — possibly linked to <b>trade wind variability</b>
          or <b>oceanic upwelling</b> effects.
        </div>
        <span class="alt-finding-tag tag-blue">Weak signal · RMSE improvement &lt; 2%</span>
      </div>

    </div>

    <!-- ── Plots + tabla de ratios ────────────────────────────────── -->
    <div class="alt-plots-grid">

      <!-- Gráfico de barras -->
      <div class="alt-plot-card">
        <div class="alt-plot-title">Historical vs. recent slope by area</div>
        <div class="alt-plot-subtitle">
          Bars extending beyond the blue (Historical OLS) indicate acceleration;
          bars falling short indicate a slowdown.
        </div>
  '))

  # ── Gráfico de barras ──────────────────────────────────────────────────
  plot_data <- acceleration %>%
    mutate(
      AREA = factor(AREA, levels = rev(area_order)),
      `Historical OLS`          = slope_hist * 100,
      `Yeo-Johnson (recent)`    = yj_recent * 100,
      `Changepoint (segment 2)` = if ("slope2_cp" %in% names(.)) slope2_cp * 12 * 100 else NA_real_
    ) %>%
    select(AREA, `Historical OLS`, `Yeo-Johnson (recent)`, `Changepoint (segment 2)`) %>%
    pivot_longer(cols = -AREA, names_to = "method", values_to = "slope") %>%
    filter(!is.na(slope))

  print(
    ggplot(plot_data, aes(x = slope, y = AREA, fill = method)) +
      geom_col(position = position_dodge(width = 0.72), width = 0.62) +
      geom_vline(xintercept = 0, linewidth = 0.4, colour = "grey60") +
      scale_fill_manual(values = c(
        "Historical OLS"          = "#1B4965",
        "Yeo-Johnson (recent)"    = "#E05C2A",
        "Changepoint (segment 2)" = "#2C6E8A"
      )) +
      labs(x = "°C/century", y = NULL, fill = NULL) +
      theme_minimal(base_size = 11, base_family = "sans") +
      theme(
        panel.grid.minor  = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y       = element_text(face = "bold", size = 10),
        legend.position   = "top",
        legend.text       = element_text(size = 10),
        plot.margin       = margin(4, 8, 4, 4)
      )
  )

  htmltools::HTML(paste0('
      </div>

      <!-- Tabla de ratios + nota RMSE -->
      <div class="alt-table-card">
        <div class="alt-table-title">Ratio: recent rate ÷ historical rate</div>
        <div class="alt-table-subtitle">
          A ratio of 1.27 means warming is 27% faster now than the long-run average.
          A ratio of 0.80 means it is 20% slower.
        </div>

        <table class="alt-ratio-table">
          <thead><tr>
            <th>Area</th>
            <th>YJ / hist.</th>
            ', header_cp, '
          </tr></thead>
          <tbody>', ratio_rows, '</tbody>
        </table>

        <div style="margin-top:10px; font-size:11.5px; color:#888">
          <span class="ratio-pill ratio-accel">accel &gt;1.05</span>
          <span class="ratio-pill ratio-stable">stable ≈1</span>
          <span class="ratio-pill ratio-decel">decel &lt;0.95</span>
        </div>

        <div class="alt-rmse-note" style="margin-top:14px">
          <b>RMSE check:</b> all alternative models improve fit by &lt;2% over OLS,
          confirming that the extra complexity is not justified. The linear model is the
          right choice for describing the full 1970–2024 period.
        </div>
      </div>

    </div><!-- /.alt-plots-grid -->
  </div><!-- /.alt-shell -->
  '))

} else {
  htmltools::HTML("<div class='caja-discusion'>Alternative models not available in this run.</div>")
}
```
